---
title: "hPTM signatures association with orphan genes"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Here, I load the supplementary table from Jeromine's hPTM signatures and examine the association between the signatures and orphan genes.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(ggrastr)
library(readxl)
```

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")
Ec_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Ddud_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Dher_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Und_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Spro_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Si_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
```

Note: Scytosiphon was misspelt in the table

```{r}
preprocess_ps_sig_orphan <- function(Species_ps){
        Species_ps.filt <- Species_ps |> 
                tidyr::drop_na() |>
                dplyr::mutate(rank = as.numeric(rank)) |>
                tidyr::drop_na(rank)
        max_ps <- max(Species_ps.filt$rank)
        if(any(colnames(Species_ps.filt) %in% c("Signature.Female", "Signature.Male"))){
                Species_ps.filt2 <- 
                        Species_ps.filt |>
                        dplyr::mutate(Signature.Female = factor(
                                Signature.Female, levels = paste0('S',1:16))) |>
                        dplyr::mutate(Signature.Male = factor(
                                Signature.Male, levels = paste0('S',1:16))) |>
                dplyr::mutate(n_all.Female = dplyr::n(), .by = Signature.Female) |>
                dplyr::mutate(n_all.Male = dplyr::n(), .by = Signature.Male)
        }else{
                Species_ps.filt2 <- 
                        Species_ps.filt |>
                        dplyr::mutate(Signature = factor(
                                Signature, levels = paste0('S',1:16))) |>
                dplyr::mutate(n_all = dplyr::n(), .by = Signature)
        }
        Species_ps.filt3 <- Species_ps.filt2 |>
                dplyr::mutate(orphan = dplyr::case_when(
                        rank == max_ps ~ "orphan",
                        .default = "nonorphan"
                ))
                # dplyr::filter(rank == max_ps)
        return(Species_ps.filt3)
}

color_table <-
        tibble::tibble(
                Signature = paste0('S',1:16),
                Colour = c("#a31d21", "#e43a30", "#f36f57", "#f8a68d", "#fddfd1", "#1b733b",
            "#2daf66", "#6ec283", "#aed9b1", "#9392c8", "#bfbfdf", "#e9eef8",
            "#2a74bb", "#61abd7", "#afd1e5", "#999999")
        )

plot_style <- function(p_in){
        p_out <-
                p_in +
                ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
                ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
                ggplot2::theme_classic() +
                ggplot2::scale_colour_manual(values = rev(color_table$Colour)) +
                ggplot2::theme(legend.position = "none") +
                ggplot2::coord_flip() +
                ggplot2::scale_x_discrete(limits=rev) +
                ggplot2::scale_y_continuous(breaks = scales::breaks_pretty())
        return(p_out)
}

plot_style_col <- function(p_in){
        p_out <-
                p_in +
                ggplot2::geom_col(width = 0.8, colour = "black") +
                ggplot2::theme_classic() +
                ggplot2::scale_fill_manual(values = rev(color_table$Colour)) +
                ggplot2::theme(legend.position = "none") +
                ggplot2::coord_flip() +
                # ggplot2::scale_x_discrete(limits=rev) +
                ggplot2::scale_y_continuous(breaks = scales::breaks_pretty())
        return(p_out)
}

sig_for_leftjoin <-
        paste0('S',1:16) |>
        factor(levels = paste0('S',1:16)) |>
        tibble::as_tibble()
```

plotting signatures

# hPTM signatures and number of genes

Ectocarpus

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ec_ps.filt_orphan.F <-
        Ec_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Female, orphan)) |>
        dplyr::distinct(Signature.Female, n_all.Female, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Ec_ps.filt_orphan.M <-
        Ec_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Male, orphan)) |>
        dplyr::distinct(Signature.Male, n_all.Male, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Ec_ps.filt_orphan.F.total <-
        sum(Ec_ps.filt_orphan.F$orphan,na.rm = TRUE)/sum(Ec_ps.filt_orphan.F$n_all.Female)

Ec_ps.filt_orphan.M.total <-
        sum(Ec_ps.filt_orphan.M$orphan,na.rm = TRUE)/sum(Ec_ps.filt_orphan.M$n_all.Male)


Ec_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Ectocarpus", subtitle = "female")

Ec_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Ectocarpus", subtitle = "male")

Ec_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = 100*orphan/n_all.Female)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Ec_ps.filt_orphan.F.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Ectocarpus", subtitle = "female")

Ec_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = 100*orphan/n_all.Male)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Ec_ps.filt_orphan.M.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Ectocarpus", subtitle = "male")
```


Desmarestia-dudresnayi

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ddud_ps.filt_orphan <-
        Ddud_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature, orphan)) |>
        dplyr::distinct(Signature, n_all, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Ddud_ps.filt_orphan.total <-
        sum(Ddud_ps.filt_orphan$orphan,na.rm = TRUE)/sum(Ddud_ps.filt_orphan$n_all)

Ddud_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, fill = Signature, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")

Ddud_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, fill = Signature, y = 100*orphan/n_all)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Ddud_ps.filt_orphan.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")
```


Desmarestia-herbacea

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Dher_ps.filt_orphan.F <-
        Dher_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Female, orphan)) |>
        dplyr::distinct(Signature.Female, n_all.Female, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Dher_ps.filt_orphan.M <-
        Dher_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Male, orphan)) |>
        dplyr::distinct(Signature.Male, n_all.Male, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Dher_ps.filt_orphan.F.total <-
        sum(Dher_ps.filt_orphan.F$orphan,na.rm = TRUE)/sum(Dher_ps.filt_orphan.F$n_all.Female)

Dher_ps.filt_orphan.M.total <-
        sum(Dher_ps.filt_orphan.M$orphan,na.rm = TRUE)/sum(Dher_ps.filt_orphan.M$n_all.Male)


Dher_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Desmarestia herbacea", subtitle = "female")

Dher_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Desmarestia herbacea", subtitle = "male")

Dher_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = 100*orphan/n_all.Female)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Dher_ps.filt_orphan.F.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Desmarestia herbacea", subtitle = "female")

Dher_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = 100*orphan/n_all.Male)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Dher_ps.filt_orphan.M.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Desmarestia herbacea", subtitle = "male")
```


Undaria

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Und_ps.filt_orphan.F <-
        Und_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Female, orphan)) |>
        dplyr::distinct(Signature.Female, n_all.Female, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Und_ps.filt_orphan.M <-
        Und_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Male, orphan)) |>
        dplyr::distinct(Signature.Male, n_all.Male, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Und_ps.filt_orphan.F.total <-
        sum(Und_ps.filt_orphan.F$orphan,na.rm = TRUE)/sum(Und_ps.filt_orphan.F$n_all.Female)

Und_ps.filt_orphan.M.total <-
        sum(Und_ps.filt_orphan.M$orphan,na.rm = TRUE)/sum(Und_ps.filt_orphan.M$n_all.Male)


Und_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Undaria", subtitle = "female")

Und_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Undaria", subtitle = "male")

Und_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = 100*orphan/n_all.Female)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Und_ps.filt_orphan.F.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Undaria", subtitle = "female")

Und_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = 100*orphan/n_all.Male)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Und_ps.filt_orphan.M.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Undaria", subtitle = "male")
```

Males removed since not all signatures are present (contains orphans).

Scytosiphon

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Spro_ps.filt_orphan.F <-
        Spro_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Female, orphan)) |>
        dplyr::distinct(Signature.Female, n_all.Female, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Spro_ps.filt_orphan.M <-
        Spro_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature.Male, orphan)) |>
        dplyr::distinct(Signature.Male, n_all.Male, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Spro_ps.filt_orphan.F.total <-
        sum(Spro_ps.filt_orphan.F$orphan)/sum(Spro_ps.filt_orphan.F$n_all.Female)

Spro_ps.filt_orphan.M.total <-
        sum(Spro_ps.filt_orphan.M$orphan)/sum(Spro_ps.filt_orphan.M$n_all.Male)


Spro_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Scytosiphon", subtitle = "female")

Spro_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Scytosiphon", subtitle = "male")

Spro_ps.filt_orphan.F |>
        ggplot2::ggplot(aes(x = Signature.Female, fill = Signature.Female, y = 100*orphan/n_all.Female)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Spro_ps.filt_orphan.F.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Scytosiphon", subtitle = "female")

Spro_ps.filt_orphan.M |>
        ggplot2::ggplot(aes(x = Signature.Male, fill = Signature.Male, y = 100*orphan/n_all.Male)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Spro_ps.filt_orphan.M.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Scytosiphon", subtitle = "male")
```


Schizocladia

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Si_ps.filt_orphan <-
        Si_ps |> 
        preprocess_ps_sig_orphan() |>
        dplyr::mutate(n_orphan = dplyr::n(), .by = c(Signature, orphan)) |>
        dplyr::distinct(Signature, n_all, n_orphan, orphan) |>
        tidyr::pivot_wider(names_from = "orphan", values_from = "n_orphan")

Si_ps.filt_orphan.total <-
        sum(Si_ps.filt_orphan$orphan)/sum(Si_ps.filt_orphan$n_all)

Si_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, fill = Signature, y = orphan)) |>
        plot_style_col() +
        ggplot2::labs(x = "signature", y = "number of orphan",
                      title = "Schizocladia", subtitle = "unknown")

Si_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, fill = Signature, y = 100*orphan/n_all)) |>
        plot_style_col() +
        ggplot2::geom_hline(yintercept = 100*Si_ps.filt_orphan.total, colour = "black", linetype = 3) +
        ggplot2::labs(x = "signature", y = "% orphan within signature",
                      title = "Schizocladia", subtitle = "unknown")
```

Is it possible to combine plots?

```{r warning=FALSE, message=FALSE}
all_ps.filt_orphan.total <-
        tibble::tibble(
                sp = c("Ec", "Spro", "Und", "Ddud", "Dher", "Si"),
                p_total = c(
                        Ec_ps.filt_orphan.F.total,
                        Spro_ps.filt_orphan.F.total,
                        Und_ps.filt_orphan.F.total,
                        Ddud_ps.filt_orphan.total,
                        Dher_ps.filt_orphan.F.total,
                        Si_ps.filt_orphan.total
                )
        )
all_ps.filt_orphan <-
        Ec_ps.filt_orphan.F |> 
        dplyr::mutate(sp = "Ec") |> 
        dplyr::rename(Signature = Signature.Female, 
                      n_all = n_all.Female) |>
        rbind(
                Und_ps.filt_orphan.F |> dplyr::mutate(sp = "Und") |> 
                        dplyr::rename(Signature = Signature.Female, 
                                      n_all = n_all.Female)
        ) |>
        rbind(
                Ddud_ps.filt_orphan |> dplyr::mutate(sp = "Ddud")
        ) |>
        rbind(
                Dher_ps.filt_orphan.F |> dplyr::mutate(sp = "Dher") |> 
                        dplyr::rename(Signature = Signature.Female, 
                                      n_all = n_all.Female)
        ) |>
        rbind(
                Spro_ps.filt_orphan.F |> dplyr::mutate(sp = "Spro") |> 
                        dplyr::rename(Signature = Signature.Female, 
                                      n_all = n_all.Female)
        ) |>
        rbind(
                Si_ps.filt_orphan |> dplyr::mutate(sp = "Si")
        ) |> 
        dplyr::left_join(all_ps.filt_orphan.total) |>
        dplyr::mutate(orphan = dplyr::case_when(
                is.na(orphan) ~ 0,
                .default = orphan
        )) |>
        dplyr::mutate(p_orphan = orphan/n_all,
                      p_diff = p_orphan - p_total)
```

```{r warning=FALSE, message=FALSE, fig.width=3, fig.height=2.5}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

all_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, colour = sp, group = sp, y = 100*p_orphan)) +
        ggplot2::geom_line(size = 1, alpha = 0.8) +
        ggplot2::labs(x = "signature", y = "% orphan within signature") +
        ggplot2::theme_classic() +
        ggplot2::coord_flip() +
        ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::scale_colour_manual(values=cbPalette)

all_ps.filt_orphan |>
        ggplot2::ggplot(aes(x = Signature, colour = sp, group = sp, y = 100*p_diff)) +
        ggplot2::geom_line(size = 1, alpha = 0.8) +
        ggplot2::labs(x = "signature", y = "% orphan within - across signature") +
        ggplot2::theme_classic() +
        ggplot2::coord_flip() +
        ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::scale_colour_manual(values=cbPalette)
```

# session info

```{r}
devtools::session_info()
```
