---
title: "Signature profile of sex chromosomes vs autosomes"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

## HMG signature

`Ec-13_001750` is the HMG gene in Ectocarpus. What is the signature and how does this differ in other species?

```
OG0000867	Ddudresnayi_g4067.t1, Ddudresnayi_g4067.t2, Ddudresnayi_g7528.t1, Ddudresnayi_g7528.t2	prot_D-herbacea_M_contig98.16175.1	Ec-13_001750.1	Upin_00015662	prot_S-ischiensis_contig127.2856.1, prot_S-ischiensis_contig86.20216.1	prot_S-promiscuus_M_contig1.143.1, prot_S-promiscuus_M_contig1.146.1
```

Let's check their signatures.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)
library(readxl)
```

Read and make tx2gene file to replace gene IDs for some species

```{r warning=FALSE, message=FALSE}
Und_tx2gene <- readr::read_tsv("data/TPM/Und_tx2gene.tsv",col_select = c(1,2), col_names = c("tx", "gene_id"))
Si_tx2gene <- readr::read_tsv("data/TPM/Schizocladia-ischiensis.tsv",col_select = c(1,2))
```

Read signature master file

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")

Ec_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Ddud_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Dher_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Spro_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))

# species where renaming is needed

Und_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        # rename gene_id
        dplyr::left_join(Und_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
Si_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        # rename gene_id
        dplyr::left_join(Si_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
```

Read and process orthogroup data to obtain one-to-one orthologues

```{r warning=FALSE, message=FALSE}
# filter longest isoform in Ectocarpus
library(rtracklayer)

Ec_data.gtf <- rtracklayer::import("data/annotation/Ec32_V5.longest_mRNA.SDR_F.AGAT.mod.gtf")
Ec_data.gtf.df <- Ec_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id) |>
        dplyr::distinct()
```

```{r warning=FALSE, message=FALSE}
og_data <-
        readr::read_tsv("data/Orthogroups.tsv") |>
        tidyr::drop_na() |>
        tidyr::pivot_longer(
                cols = -Orthogroup,
                names_to = "species",
                values_to = "gene_id"
                ) |>
        dplyr::filter(!is.na(gene_id)) |>
        tidyr::separate_rows(gene_id, sep = ",\\s*") |>
        dplyr::filter(gene_id != "") |>
        # filter isoforms
        dplyr::filter(!stringr::str_detect(gene_id, "\\.t[0-9]+$") | stringr::str_detect(gene_id, "\\.t1$")) |>
        dplyr::filter(
                (species == "PUBLIC_Ectocarpus-sp7_proteins" & gene_id %in% Ec_data.gtf.df$gene_id) |
                        (species != "PUBLIC_Ectocarpus-sp7_proteins")
        )
                
og_data2 <- 
        og_data |>
        dplyr::mutate(
                gene_id = stringr::str_remove(gene_id, "\\.t1$"),
                gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_"),
                species = dplyr::case_when(
                        stringr::str_detect(species,"dudresnayi") ~ "Ddud",
                        stringr::str_detect(species,"herbacea") ~ "Dher",
                        stringr::str_detect(species,"Ectocarpus") ~ "Ec",
                        stringr::str_detect(species,"Undaria") ~ "Und",
                        stringr::str_detect(species,"Scytosiphon") ~ "Spro",
                        stringr::str_detect(species,"Schizocladia") ~ "Si"
                ))
```

```{r warning=FALSE, message=FALSE}
sig.all <- 
        Ec_sig |> dplyr::select(1,Signature.Female,Signature.Male) |>
        dplyr::bind_rows(Ddud_sig) |>
        dplyr::bind_rows(Dher_sig |> dplyr::select(1,Signature.Female,Signature.Male)) |>
        dplyr::bind_rows(Spro_sig |> dplyr::select(1,Signature.Female,Signature.Male)) |>
        dplyr::bind_rows(Und_sig |> dplyr::select(1,Signature.Female,Signature.Male)) |>
        dplyr::bind_rows(Si_sig)


og_data_sig <- 
        og_data2 |> dplyr::filter(Orthogroup == "OG0000867") |>
        dplyr::left_join(sig.all) |>
        dplyr::select(-Orthogroup) |>
        dplyr::distinct()

og_data_sig[og_data_sig == "NA"] <- NA
```

```{r warning=FALSE, message=FALSE}
og_data_sig |>
        gt::gt()
```

Nothing so different to what we thought...


## Signature difference between autosomes and sex chromosomes


```{r warning=FALSE, message=FALSE}
Ec_sig.chr <- 
        Ec_sig |>
        dplyr::left_join(Ec_data.gtf.df) |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "chr_13" ~ "sex",
                .default = "autosome"
        )) |>
        dplyr::mutate(Signature.Female)

Ec_sig.chr[Ec_sig.chr == "NA"] <- NA

color_table <-
        tibble::tibble(
                Signature = paste0('S',1:16),
                Colour = c("#a31d21", "#e43a30", "#f36f57", "#f8a68d", "#fddfd1", "#1b733b",
            "#2daf66", "#6ec283", "#aed9b1", "#9392c8", "#bfbfdf", "#e9eef8",
            "#2a74bb", "#61abd7", "#afd1e5", "#999999")
        )
```

In Ectocarpus.

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=5}
Ec_sig.chr |>
        dplyr::mutate(Signature.Female = factor(Signature.Female, levels = paste0('S',16:1))) |>
        tidyr::drop_na() |>
        ggplot2::ggplot(aes(x = chr_type, fill = Signature.Female)) +
        ggplot2::geom_bar(position="fill") +
        ggplot2::scale_y_continuous(labels = scales::percent) +
        ggplot2::scale_fill_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "Ectocarpus signature on chr_13",
             x = "Chromosome type",
             y = "Proportion of signatures",
             fill = "Signature") +
        ggplot2::theme_classic()

Ec_sig.chr |>
        dplyr::count(.by = chr_type)
```


And for other species.

```{r warning=FALSE, message=FALSE}
# Spro chr_13
Spro_data.gtf <- rtracklayer::import("data/annotation/convert_gffread/Scytosiphon-promiscuus_MALE.chr.gffread.gff")
Spro_data.gtf.df <- Spro_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = ID) |>
        dplyr::distinct() |>
        dplyr::mutate(sp = "Spro") |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "chr_13" ~ "sex",
                .default = "autosome"
        ))
# Und HiC_scaffold_23
# there are apparently 30 'chromosome-level' scaffolds.
# I confirmed this with seqkit.
Und_data.gtf <- rtracklayer::import("data/annotation/PUBLIC_Undaria-pinnatifida_MALE.jv3.gtf")
Und_data.gtf.df <- Und_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = transcript_id) |>
        dplyr::distinct() |>
        dplyr::mutate(sp = "Und") |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "HiC_scaffold_23" ~ "sex",
                .default = "autosome"
        ))
# Dher chr_03
Dher_data.gtf <- rtracklayer::import("data/annotation/convert_gffread/Desmarestia-herbacea.reference.gt.AGAT.gffread.gff")
Dher_data.gtf.df <- Dher_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = ID) |>
        dplyr::distinct() |>
        dplyr::mutate(sp = "Dher") |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "chr_03" ~ "sex",
                .default = "autosome"
        ))
# Ddud pseudo-chr_03
Ddud_data.gtf <- rtracklayer::import("data/annotation/Desmarestia-dudresnayi.cleaned2025.jv.gffread.gtf")
Ddud_data.gtf.df <- Ddud_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = transcript_id) |>
        dplyr::distinct() |>
        dplyr::mutate(sp = "Ddud") |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "chr_03" ~ "pseudo-sex",
                .default = "autosome"
        )) |>
        dplyr::mutate(gene_id = stringr::str_remove(gene_id, "\\.t1$")) # remove isoform suffix

gtf.df.all <-
        Ec_data.gtf.df |>
        dplyr::mutate(sp = "Ec") |>
        dplyr::mutate(chr_type = dplyr::case_when(
                seqnames == "chr_13" ~ "sex",
                .default = "autosome"
        )) |>
        rbind(Spro_data.gtf.df) |>
        rbind(Und_data.gtf.df) |>
        rbind(Dher_data.gtf.df) |>
        rbind(Ddud_data.gtf.df)

sig.all <- 
        Ec_sig |> dplyr::select(1,Signature = Signature.Female) |>
        rbind(Ddud_sig) |>
        rbind(Dher_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Spro_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Und_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Si_sig)

sig.all[sig.all == "NA"] <- NA  # replace "NA" with NA

sig.all.gtf.df.all <- 
        sig.all |> 
        dplyr::left_join(gtf.df.all) |>
        dplyr::mutate(Signature = factor(Signature, levels = paste0('S',16:1)))

rect_df <- sig.all.gtf.df.all |>
        dplyr::distinct(sp) |>
        dplyr::filter(sp == "Ddud") |>
        dplyr::mutate(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Spro", "Und", "Dher", "Ddud"
        )))
```

```{r warning=FALSE, message=FALSE, fig.width=6, fig.height=5}
p_sig.all.gtf.df.all <- sig.all.gtf.df.all |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Spro", "Und", "Dher", "Ddud"
        ))) |>
        tidyr::drop_na()

p_sig.all.gtf.df.all |>
        ggplot2::ggplot(aes(x = chr_type, fill = Signature)) +
          # shading for monoicous species only
          ggplot2::geom_rect(
            data = rect_df,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE,
            fill = "grey", alpha = 0.2
          ) +
          # the rest
        ggplot2::geom_bar(position="fill") +
        ggplot2::scale_y_continuous(labels = scales::percent) +
        ggplot2::scale_fill_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "All species",
             x = "Chromosome type",
             y = "Proportion of signatures",
             fill = "Signature") +
        ggplot2::theme_classic() +
        ggplot2::facet_grid(.~ sp, scales = "free_x") +
        ggplot2::guides(x = guide_axis(angle = 90))


sig.all.gtf.df.all |>
        tidyr::drop_na() |>
        dplyr::summarise(n_genes = n(), .by = c(sp, chr_type))
```



```{r warning=FALSE, message=FALSE, fig.width=6, fig.height=3}
p_sig.all.gtf.df.all |>
  ggplot2::ggplot(aes(x = chr_type, fill = Signature)) +
    # shading
    ggplot2::geom_rect(
      data = rect_df,
      aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
      inherit.aes = FALSE,
      fill = "grey", alpha = 0.3
    ) +
    ggplot2::geom_bar(position="fill") +
    ggplot2::scale_y_continuous(labels = scales::percent) +
    ggplot2::scale_fill_manual(values = setNames(color_table$Colour, color_table$sp)) +
    ggplot2::labs(
      title = "All species",
      x = "Chromosome type",
      y = "Proportion of signatures",
      fill = "Signature"
    ) +
    ggplot2::theme_classic() +
    ggplot2::facet_grid(. ~ sp, scales = "free_x") +
    ggplot2::guides(x = guide_axis(angle = 90),
                    fill=guide_legend(ncol=2))
```