---
title: "hPTM signatures association with gene age and expression dynamics"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Here, I load the supplementary table from Jeromine's hPTM signatures and examine the association between the signatures and evolutionary (as captured by gene age) and expression dynamics (as captured by tau)

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(ggrastr)
library(readxl)
library(scales)
```

We use ggrastr because it helps with plotting large number of points.

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")
Ec_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Ddud_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Dher_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Und_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Spro_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Si_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
```

Note: Scytosiphon was misspelt in the table

```{r}
preprocess_ps_sig <- function(Species_ps){
        Species_ps.filt <- Species_ps |> 
                tidyr::drop_na() |>
                dplyr::mutate(rank = as.numeric(rank)) |>
                tidyr::drop_na(rank)
        if(any(colnames(Species_ps.filt) %in% c("Signature.Female", "Signature.Male"))){
                Species_ps.filt2 <- 
                        Species_ps.filt |>
                        dplyr::mutate(Signature.Female = factor(
                                Signature.Female, levels = paste0('S',1:16))) |>
                        dplyr::mutate(Signature.Male = factor(
                                Signature.Male, levels = paste0('S',1:16)))
        }else{
                Species_ps.filt2 <- 
                        Species_ps.filt |>
                        dplyr::mutate(Signature = factor(
                                Signature, levels = paste0('S',1:16)))
        }
        return(Species_ps.filt2)
}

color_table <-
        tibble::tibble(
                Signature = paste0('S',1:16),
                Colour = c("#a31d21", "#e43a30", "#f36f57", "#f8a68d", "#fddfd1", "#1b733b",
            "#2daf66", "#6ec283", "#aed9b1", "#9392c8", "#bfbfdf", "#e9eef8",
            "#2a74bb", "#61abd7", "#afd1e5", "#999999")
        )

plot_style <- function(p_in){
        p_out <-
                p_in +
                ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
                ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
                ggplot2::theme_classic() +
                ggplot2::scale_colour_manual(values = rev(color_table$Colour)) +
                ggplot2::theme(legend.position = "none") +
                ggplot2::coord_flip() +
                # ggplot2::scale_x_discrete(limits=rev) +
                ggplot2::scale_y_continuous(breaks = scales::breaks_pretty())
        return(p_out)
}
```

plotting signatures

# hPTM signatures and mean expression or gene age.

Ectocarpus

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ec_ps.filt <-
        Ec_ps |> 
        preprocess_ps_sig()
Ec_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank",
                      title = "Ectocarpus", subtitle = "female")
Ec_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank",
                      title = "Ectocarpus", subtitle = "male")
# expression
Ec_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = log2_TPM_plus_1.Female)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)",
                      title = "Ectocarpus", subtitle = "female")
Ec_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = log2_TPM_plus_1.Male)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)",
                      title = "Ectocarpus", subtitle = "male")
```

```{r}
kruskal.test(log2_TPM_plus_1.Female ~ Signature.Female, data = Ec_ps.filt)
kruskal.test(log2_TPM_plus_1.Male ~ Signature.Male, data = Ec_ps.filt)
kruskal.test(rank ~ Signature.Female, data = Ec_ps.filt)
kruskal.test(rank ~ Signature.Male, data = Ec_ps.filt)
```

In terms of information, the grouping reduces entropy w.r.t. gene age ranks.

Desmarestia-dudresnayi

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ddud_ps.filt <-
        Ddud_ps |> 
        preprocess_ps_sig()
Ddud_ps.filt |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank", 
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")
# expression
Ddud_ps.filt |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = log2_TPM_plus_1)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)", 
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")
```

```{r}
kruskal.test(log2_TPM_plus_1 ~ Signature, data = Ddud_ps.filt)
kruskal.test(rank ~ Signature, data = Ddud_ps.filt)
```

Desmarestia-herbacea

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Dher_ps.filt <-
        Dher_ps |> 
        preprocess_ps_sig()
Dher_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank",  
                      title = "Desmarestia herbacea", subtitle = "female")
Dher_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank", 
                      title = "Desmarestia herbacea", subtitle = "male")
# expression
Dher_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = log2_TPM_plus_1.Female)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)", 
                      title = "Desmarestia herbacea", subtitle = "female")
Dher_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = log2_TPM_plus_1.Male)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)",  
                      title = "Desmarestia herbacea", subtitle = "male")
```

```{r}
kruskal.test(log2_TPM_plus_1.Female ~ Signature.Female, data = Dher_ps.filt)
kruskal.test(log2_TPM_plus_1.Male ~ Signature.Male, data = Dher_ps.filt)
kruskal.test(rank ~ Signature.Female, data = Dher_ps.filt)
kruskal.test(rank ~ Signature.Male, data = Dher_ps.filt)
```

Undaria

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Und_ps.filt <-
        Und_ps |> 
        preprocess_ps_sig()
Und_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank", 
                      title = "Undaria", subtitle = "female")
Und_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank", 
                      title = "Undaria", subtitle = "male")
# expression
Und_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = log2_TPM_plus_1.Female)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)", 
                      title = "Undaria", subtitle = "female")
Und_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = log2_TPM_plus_1.Male)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)", 
                      title = "Undaria", subtitle = "male")
```

```{r}
kruskal.test(log2_TPM_plus_1.Female ~ Signature.Female, data = Und_ps.filt)
kruskal.test(log2_TPM_plus_1.Male ~ Signature.Male, data = Und_ps.filt)
kruskal.test(rank ~ Signature.Female, data = Und_ps.filt)
kruskal.test(rank ~ Signature.Male, data = Und_ps.filt)
```

Scytosiphon

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Spro_ps.filt <-
        Spro_ps |> 
        preprocess_ps_sig()
Spro_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank",
                      title = "Scytosiphon", subtitle = "female")
Spro_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank",
                      title = "Scytosiphon", subtitle = "male")
# expression
Spro_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = log2_TPM_plus_1.Female)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)",
                      title = "Scytosiphon", subtitle = "female")
Spro_ps.filt |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = log2_TPM_plus_1.Male)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)",
                      title = "Scytosiphon", subtitle = "male")
```

```{r}
kruskal.test(log2_TPM_plus_1.Female ~ Signature.Female, data = Spro_ps.filt)
kruskal.test(log2_TPM_plus_1.Male ~ Signature.Male, data = Spro_ps.filt)
kruskal.test(rank ~ Signature.Female, data = Spro_ps.filt)
kruskal.test(rank ~ Signature.Male, data = Spro_ps.filt)
```

Schizocladia

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Si_ps.filt <-
        Si_ps |> 
        preprocess_ps_sig()
Si_ps.filt |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = rank)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "phylorank", 
                      title = "Schizocladia", subtitle = "unknown")
# expression
Si_ps.filt |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = log2_TPM_plus_1)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "log(TPM+1)", 
                      title = "Schizocladia", subtitle = "unknown")
```


```{r}
kruskal.test(log2_TPM_plus_1 ~ Signature, data = Si_ps.filt)
kruskal.test(rank ~ Signature, data = Si_ps.filt)
```

Across all species, the MI is around 4 to 8%.

# hPTM signatures and expression variability.

## using the hourglass data

using Ectocarpus data from hourglass

```{r warning=FALSE, message=FALSE}
library(tidyr); library(duckplyr)
options(duckdb.materialize_message = FALSE)
```


```{r warning=FALSE, message=FALSE}
Ec_TPM <- readr::read_csv("data/hourglass_data/Ec_abundance.csv") |> 
        # dplyr::mutate(GeneID = stringr::str_remove(gene_id, "\\..$")) |>
        tibble::column_to_rownames("GeneID")
```

```{r}
# check Kryuchkova-Mostacci & Robinson-Rechavi for more functions to add.
specificity <- function(abundance, measure = "tau"){
        abundance_longer <- 
                abundance |>
                duckplyr::as_duckplyr_tibble() |>
                duckplyr::select(1, where(is.numeric)) |>
                tidyr::pivot_longer(!1, names_to = "sample", values_to = "TPM")
        if(measure == "gini"){
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                gini = DescTools::Gini(TPM),
                                .by = GeneID
                        )
        } else if(measure == "tau"){
                n_col <- ncol(abundance) - 1
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                tau = sum(1 - (TPM/max(TPM)))/(n_col-1),
                                .by = GeneID
                        )
        } else if(measure == "H"){
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                H = DescTools::Entropy(TPM/sum(TPM)),
                                .by = GeneID
                        )
        } else if(measure == "rH"){
                # KL divergence from uniform distribution
                n_col <- ncol(abundance) - 1
                x_1 <- 
                breadth_out <- abundance_longer |>
                        duckplyr::mutate(
                                # rH = entropy::KL.empirical(y1 = TPM/sum(TPM), y2 = 1/n_col),
                                y1 = TPM/sum(TPM),
                                y2 = 1/n_col,
                                .by = GeneID
                        ) |>
                        duckplyr::summarise(
                                rH = entropy::KL.empirical(y1, y2),
                                .by = GeneID
                        )
        } else if(measure == "mean"){
                # not really a specificity but useful nonetheless
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                meanTPM = mean(TPM),
                                .by = GeneID
                        )
        } else {
                stop("Specify measure: \"gini\", \"tau\", \"H\", \"rH\", \"mean\"")
        }
        return(breadth_out)
}
```

```{r message=FALSE, warning=FALSE, results='hide'}
Ec_TPM.log2 <- log2(Ec_TPM + 1) |> tibble::rownames_to_column("GeneID")
# Ec_TPM.log2 <- log2(Ec_TPM[,c(7:10)] + 1) |> tibble::rownames_to_column("GeneID")

Ec_tau.log2 <- specificity(Ec_TPM.log2, measure = "tau")
Ec_H.log2 <- specificity(Ec_TPM.log2, measure = "H")
Ec_rH.log2 <- specificity(Ec_TPM.log2, measure = "rH")
```

```{r message=FALSE, warning=FALSE, results='hide'}
Ec_combined <- 
        Ec_ps.filt |> 
        # dplyr::mutate(GeneID = stringr::str_remove(gene_id, "\\..$")) |>
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(Ec_H.log2) |>
        dplyr::inner_join(Ec_tau.log2) |>
        dplyr::inner_join(Ec_rH.log2)
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Ectocarpus", subtitle = "female")
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Ectocarpus", subtitle = "male")
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = H)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "H (bits)",
                      title = "Ectocarpus", subtitle = "female")
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = H)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "H (bits)",
                      title = "Ectocarpus", subtitle = "male")
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Ectocarpus", subtitle = "female")
Ec_combined |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Ectocarpus", subtitle = "male")
```

It should be noted in these plots that tau, H and rH were computed using all data (not just restricted to just males or females).

## using Jero's RNA-seq data

```{r message=FALSE, warning=FALSE, results='hide'}
selectGenes <- function(counts, min.count=3){
  keep <-
    counts[rowMeans(counts)>min.count, ]
  return(keep)
}
# this is necessary because if the mean expression is 0, then we cannot compute the variability measures.
# also, if mean expression is low, it might be noisy.
```

```{r message=FALSE, warning=FALSE, results='hide'}
Ec_TPM.j <- readr::read_tsv("data/TPM/Ectocarpus-sp7-WT.tsv") |>
        dplyr::select(GeneID = gene_id, where(is.numeric)) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
Ddud_TPM.j <- readr::read_tsv("data/TPM/Desmarestia-dudresnayi.tsv") |>
        dplyr::select(GeneID = gene_id, where(is.numeric)) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
Dher_TPM.j <- readr::read_tsv("data/TPM/Desmarestia-herbacea.tsv") |>
        dplyr::select(GeneID = tx, where(is.numeric)) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
Und_TPM.j <- readr::read_tsv("data/TPM/Undaria-pinnatifida.tsv") |>
        dplyr::select(GeneID = gene_id, where(is.numeric)) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
Spro_TPM.j <- readr::read_tsv("data/TPM/Scytosiphon-promiscuus.tsv") |>
        dplyr::select(GeneID = tx, where(is.numeric)) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
Si_TPM.j <- readr::read_tsv("data/TPM/Schizocladia-ischiensis.tsv") |>
        dplyr::select(GeneID = gene_id, where(is.numeric)) |>
        dplyr::select(!dplyr::contains("RAP10_")) |>
        tibble::column_to_rownames("GeneID") |>
        selectGenes()
```


```{r message=FALSE, warning=FALSE, results='hide'}
Ec_TPM.j.log2 <- log2(Ec_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
Ddud_TPM.j.log2 <- log2(Ddud_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
Dher_TPM.j.log2 <- log2(Dher_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
Und_TPM.j.log2 <- log2(Und_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
Spro_TPM.j.log2 <- log2(Spro_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
Si_TPM.j.log2 <- log2(Si_TPM.j + 1) |> tibble::rownames_to_column("GeneID")
```

```{r message=FALSE, warning=FALSE, results='hide'}
Ec_combined.j <- 
        Ec_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Ec_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Ec_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Ec_TPM.j.log2, measure = "rH"))
Ddud_combined.j <- 
        Ddud_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Ddud_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Ddud_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Ddud_TPM.j.log2, measure = "rH"))
Dher_combined.j <- 
        Dher_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Dher_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Dher_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Dher_TPM.j.log2, measure = "rH"))
Und_combined.j <- 
        Und_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Und_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Und_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Und_TPM.j.log2, measure = "rH"))
Spro_combined.j <- 
        Spro_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Spro_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Spro_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Spro_TPM.j.log2, measure = "rH"))
Si_combined.j <- 
        Si_ps.filt |> 
        dplyr::rename(GeneID = gene_id) |>
        dplyr::inner_join(specificity(Si_TPM.j.log2, measure = "tau")) |>
        dplyr::inner_join(specificity(Si_TPM.j.log2, measure = "H")) |>
        dplyr::inner_join(specificity(Si_TPM.j.log2, measure = "rH"))
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ec_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Ectocarpus", subtitle = "female")
Ec_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Ectocarpus", subtitle = "male")
Ec_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Ectocarpus", subtitle = "female")
Ec_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Ectocarpus", subtitle = "male")
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Ddud_combined.j |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")

Ddud_combined.j |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Desmarestia dudresnayi", subtitle = "hermaphrodite")
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Dher_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Desmarestia herbacea", subtitle = "female")
Dher_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Desmarestia herbacea", subtitle = "male")

Dher_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Desmarestia herbacea", subtitle = "female")
Dher_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Desmarestia herbacea", subtitle = "male")
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Und_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Undaria", subtitle = "female")
Und_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Undaria", subtitle = "male")

Und_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Undaria", subtitle = "female")
Und_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Undaria", subtitle = "male")
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Spro_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Scytosiphon promiscuus", subtitle = "female")
Spro_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Scytosiphon promiscuus", subtitle = "male")

Spro_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Female, colour = Signature.Female, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Scytosiphon promiscuus", subtitle = "female")
Spro_combined.j |>
        ggplot2::ggplot(aes(x = Signature.Male, colour = Signature.Male, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Scytosiphon promiscuus", subtitle = "male")
```

```{r warning=FALSE, message=FALSE, fig.width=2.5, fig.height=3}
Si_combined.j |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = tau)) |>
        plot_style() +
        ggplot2::labs(x = "signature", y = "tau",
                      title = "Schizocladia", subtitle = "unknown")
Si_combined.j |>
        ggplot2::ggplot(aes(x = Signature, colour = Signature, y = rH)) |>
        plot_style() +
        ggplot2::scale_y_log10(labels = label_log()) +
        ggplot2::labs(x = "signature", y = "relative H (bits)",
                      title = "Schizocladia", subtitle = "unknown")
```

It should be noted again that tau and rH were computed using all data (not just restricted to just males or females).

```{r}
# Ec
kruskal.test(rH ~ Signature.Female, data = Ec_combined.j)
# Ddud
kruskal.test(rH ~ Signature, data = Ddud_combined.j)
# Dher
kruskal.test(rH ~ Signature.Female, data = Dher_combined.j)
# Und
kruskal.test(rH ~ Signature.Female, data = Und_combined.j)
# Spro
kruskal.test(rH ~ Signature.Female, data = Spro_combined.j)
# Si
kruskal.test(rH ~ Signature, data = Si_combined.j)
```

```{r warning=FALSE, message=FALSE}
all_combined.j <- 
        Dher_combined.j |> 
        dplyr::select(Signature = Signature.Female, rank, rH) |> 
        dplyr::mutate(sp = "Dher") |>
        rbind(Ddud_combined.j |> 
                      dplyr::select(Signature, rank, rH) |> 
                      dplyr::mutate(sp = "Ddud")
        ) |>
        rbind(Si_combined.j |> 
                      dplyr::select(Signature, rank, rH) |> 
                      dplyr::mutate(sp = "Si")
        ) |>
        rbind(Spro_combined.j |> 
                      dplyr::select(Signature = Signature.Female, rank, rH) |> 
                      dplyr::mutate(sp = "Spro")
        ) |>
        rbind(Ec_combined.j |> 
                      dplyr::select(Signature = Signature.Female, rank, rH) |> 
                      dplyr::mutate(sp = "Ec")
        ) |>
        rbind(Und_combined.j|> 
                      dplyr::select(Signature = Signature.Female, rank, rH) |> 
                      dplyr::mutate(sp = "Und")
        ) |>
        dplyr::summarise(
                median_rank = median(rank),
                median_rH = median(rH),
                mean_rH = mean(rH),
                .by = c(sp, Signature)
        )
```

```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=2.5}
cbPalette <- c("#999999", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

all_combined.j |>
        ggplot2::ggplot(aes(x = Signature, y = median_rank, 
                            colour = sp, group = sp)) +
        ggplot2::geom_line(size = 1, alpha = 0.8) +
        ggplot2::labs(x = "signature", y = "median phylorank", colour = "species") +
        ggplot2::theme_classic() +
        ggplot2::coord_flip() +
        ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_continuous(breaks = c(1,5,10)) +
        ggplot2::scale_colour_manual(values=cbPalette)

all_combined.j |>
        ggplot2::ggplot(aes(x = Signature, y = median_rH, 
                            colour = sp, group = sp)) +
        ggplot2::geom_line(size = 1, alpha = 0.8) +
        ggplot2::labs(x = "signature", y = "median rH (bits)", colour = "species") +
        ggplot2::theme_classic() +
        ggplot2::coord_flip() +
        ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_log10() +
        # ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::scale_colour_manual(values=cbPalette)
```


# session info

```{r}
devtools::session_info()
```
