---
title: "Signature simulation"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Since there were a few challenges to making a synteny plot for the conserved genes, we wanted to perform permutations to show that the number of orthologues with the same signature is actually significant = conserved (at least in regulatory functions).

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels)
library(readxl)
```

Read and make tx2gene file to replace gene IDs for some species.

```{r warning=FALSE, message=FALSE}
Und_tx2gene <- readr::read_tsv("data/TPM/Und_tx2gene.tsv",col_select = c(1,2), col_names = c("tx", "gene_id"))
Si_tx2gene <- readr::read_tsv("data/TPM/Schizocladia-ischiensis.tsv",col_select = c(1,2))
```

Read signature master file

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")

Ec_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Ddud_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Dher_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Spro_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))

# species where renaming is needed

Und_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        # rename gene_id
        dplyr::left_join(Und_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
Si_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        # rename gene_id
        dplyr::left_join(Si_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
```

Read and process orthogroup data to obtain one-to-one orthologues

Get a single isoform per gene (post-hoc).


```{r warning=FALSE, message=FALSE}
# filter longest isoform in Ectocarpus
library(rtracklayer)

Ec_data.gtf <- rtracklayer::import("data/annotation/Ec32_V5.longest_mRNA.SDR_F.AGAT.mod.gtf")
Ec_data.gtf.df <- Ec_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id) |>
        dplyr::distinct()
```

```{r warning=FALSE, message=FALSE}
og_data <-
        readr::read_tsv("data/Orthogroups.tsv") |>
        tidyr::drop_na() |>
        tidyr::pivot_longer(
                cols = -Orthogroup,
                names_to = "species",
                values_to = "gene_id"
                ) |>
        dplyr::filter(!is.na(gene_id)) |>
        tidyr::separate_rows(gene_id, sep = ",\\s*") |>
        dplyr::filter(gene_id != "") |>
        # filter isoforms
        dplyr::filter(!stringr::str_detect(gene_id, "\\.t[0-9]+$") | stringr::str_detect(gene_id, "\\.t1$")) |>
        dplyr::filter(
                (species == "PUBLIC_Ectocarpus-sp7_proteins" & gene_id %in% Ec_data.gtf.df$gene_id) |
                        (species != "PUBLIC_Ectocarpus-sp7_proteins")
        )

# get single copy data
og_sc.pre <- 
        og_data |>
        dplyr::summarise(n = dplyr::n(), .by = c(Orthogroup, species)) |>
        dplyr::filter(sum(n) == 6, .by = c(Orthogroup)) |>
        dplyr::distinct(Orthogroup)
og_sc <-
        og_data |>
        dplyr::filter(Orthogroup %in% og_sc.pre$Orthogroup)
og_sc.edit <-
        og_sc |>
        dplyr::mutate(
                gene_id = stringr::str_remove(gene_id, "\\.t1$"),
                gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_"),
                species = dplyr::case_when(
                        stringr::str_detect(species,"dudresnayi") ~ "Ddud",
                        stringr::str_detect(species,"herbacea") ~ "Dher",
                        stringr::str_detect(species,"Ectocarpus") ~ "Ec",
                        stringr::str_detect(species,"Undaria") ~ "Und",
                        stringr::str_detect(species,"Scytosiphon") ~ "Spro",
                        stringr::str_detect(species,"Schizocladia") ~ "Si"
                ))
```

Combine orthologue and signature data

```{r warning=FALSE, message=FALSE}
sig.all <- 
        Ec_sig |> dplyr::select(1,Signature = Signature.Female) |>
        rbind(Ddud_sig) |>
        rbind(Dher_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Spro_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Und_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Si_sig)

sig.all[sig.all == "NA"] <- NA  # replace "NA" with NA

og_sc_sig <- 
        og_sc.edit |>
        dplyr::left_join(sig.all) |>
        dplyr::distinct(Orthogroup, species, .keep_all = TRUE) |>  # keep first match only
        tidyr::pivot_wider(!gene_id, names_from = species, values_from = Signature) |>
        tidyr::drop_na()
```

Obtain empirical values for the conservation of orthologue signatures

```{r warning=FALSE, message=FALSE}
og_sc_sig.6 <- og_sc_sig |> 
        dplyr::mutate(all_same = n_distinct(c_across(Ddud:Spro)) == 1,
                      .by = Orthogroup)

og_sc_sig.6_2 <- og_sc_sig |> 
        dplyr::mutate(all_same = n_distinct(c_across(Ddud:Spro)) == 2,
                      .by = Orthogroup)

og_sc_sig.5 <- og_sc_sig |> 
        dplyr::mutate(all_same = n_distinct(c(Ddud, Dher, Ec, Und, Spro)) == 1,
                      .by = Orthogroup)

# exact conservation
table(og_sc_sig.6$all_same)

# exact conservation excluding schizocladia
table(og_sc_sig.5$all_same)

# at least 5 out of 6 species are conserved
table(og_sc_sig.6_2$all_same)
```

`294` (out of 3770) orthologues share signatures out of 6 species.
If we allow at least one species to differ then, we get `1257` conserved.
`612` orthologues share signatures out of 5 species, after excluding the outgroup Schizocladia.

Now the question is how this differs from the null hypothesis of *no conservation*.
We can permute the signatures of the orthologues and see how many orthologues have the same signature in each permutation.

If we permute only Spro data to compare conservation - so only two species.

```{r warning=FALSE, message=FALSE}
og_sc_sig.rand.2 <- 
        og_sc_sig |>
        # rsample::permutations(permute = Ddud:Spro,
        #                       times = 100) |>
        rsample::permutations(permute = Spro,
                              times = 100,
                              apparent = TRUE)

# og_sc_sig.rand$splits[[1]] %>% analysis()


count_same_fast_Spro <- function(df) {
  same <- apply(df |> dplyr::select(Ec, Spro), 1, function(x) length(unique(x)) == 1)
  sum(same)
}

og_sc_sig.rand.res.2 <- og_sc_sig.rand.2 |>
        dplyr::mutate(n_same = map_int(splits, ~ count_same_fast_Spro(analysis(.x)))) |>
        dplyr::select(id, n_same) |> 
        dplyr::mutate(data = dplyr::case_when(
                id == "Apparent" ~ "empirical",
                .default = "permutation"
        )) 
```

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
og_sc_sig.rand.res.2 |> 
        ggplot2::ggplot(aes(n_same, fill = data)) +
        ggplot2::geom_histogram(colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::labs(
                x = "orthologue w/ conserved signature",
                title = "Ec vs Spro"
        )
```

and for the furthest part of the tree Ectocarpus (Ec) vs Schizocladia (Si)

```{r warning=FALSE, message=FALSE}
og_sc_sig.rand.2Si <- 
        og_sc_sig |>
        # rsample::permutations(permute = Ddud:Spro,
        #                       times = 100) |>
        rsample::permutations(permute = Si,
                              times = 100,
                              apparent = TRUE)

# og_sc_sig.rand$splits[[1]] %>% analysis()


count_same_fast_Si <- function(df) {
  same <- apply(df |> dplyr::select(Ec, Si), 1, function(x) length(unique(x)) == 1)
  sum(same)
}

og_sc_sig.rand.res.2Si <- og_sc_sig.rand.2Si |>
        dplyr::mutate(n_same = map_int(splits, ~ count_same_fast_Si(analysis(.x)))) |>
        dplyr::select(id, n_same) |> 
        dplyr::mutate(data = dplyr::case_when(
                id == "Apparent" ~ "empirical",
                .default = "permutation"
        )) 
```

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
og_sc_sig.rand.res.2Si |> 
        ggplot2::ggplot(aes(n_same, fill = data)) +
        ggplot2::geom_histogram(colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::labs(
                x = "orthologue w/ conserved signature",
                title = "Ec vs Si"
        )
```

and between hermaphrodite and sexual Dher and Ddud

```{r warning=FALSE, message=FALSE}
og_sc_sig.rand.2herm <- 
        og_sc_sig |>
        # rsample::permutations(permute = Ddud:Spro,
        #                       times = 100) |>
        rsample::permutations(permute = Dher,
                              times = 100,
                              apparent = TRUE)

# og_sc_sig.rand$splits[[1]] %>% analysis()


count_same_fast_herm <- function(df) {
  same <- apply(df |> dplyr::select(Dher, Ddud), 1, function(x) length(unique(x)) == 1)
  sum(same)
}

og_sc_sig.rand.res.2herm <- og_sc_sig.rand.2herm |>
        dplyr::mutate(n_same = map_int(splits, ~ count_same_fast_herm(analysis(.x)))) |>
        dplyr::select(id, n_same) |> 
        dplyr::mutate(data = dplyr::case_when(
                id == "Apparent" ~ "empirical",
                .default = "permutation"
        )) 
```

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
og_sc_sig.rand.res.2herm |> 
        ggplot2::ggplot(aes(n_same, fill = data)) +
        ggplot2::geom_histogram(colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::labs(
                x = "orthologue w/ conserved signature",
                title = "Dher vs Ddud"
        )
```

Check for normality in the permutations

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
# Ec vs Spro
og_sc_sig.rand.res.2.temp <- 
        og_sc_sig.rand.res.2 |>
        dplyr::filter(data == "permutation") 
og_sc_sig.rand.res.2.temp |> 
        ggpubr::ggqqplot(x = "n_same")
shapiro.test(og_sc_sig.rand.res.2.temp$n_same)

# Ec vs Si
og_sc_sig.rand.res.2Si.temp <- 
        og_sc_sig.rand.res.2Si |>
        dplyr::filter(data == "permutation") 
og_sc_sig.rand.res.2Si.temp |> 
        ggpubr::ggqqplot(x = "n_same")
shapiro.test(og_sc_sig.rand.res.2Si.temp$n_same)

# Dher vs Ddud
og_sc_sig.rand.res.2herm.temp <- 
        og_sc_sig.rand.res.2herm |>
        dplyr::filter(data == "permutation") 
og_sc_sig.rand.res.2herm.temp |> 
        ggpubr::ggqqplot(x = "n_same")
shapiro.test(og_sc_sig.rand.res.2herm.temp$n_same)
```

```{r}
rm(og_sc_sig.rand.res.2.temp, og_sc_sig.rand.res.2Si.temp, og_sc_sig.rand.res.2herm.temp)
rm(og_sc_sig.rand.res.2, og_sc_sig.rand.res.2Si, og_sc_sig.rand.res.2herm)
```

There is a strong relationship between the evolutionary distance and the significance of the difference between the observed and permuted values.

### Permutations across all species (pairwise against Ectocarpus)

Let's do this more systematically. We want to reshuffle all species and compare the difference w.r.t. Ectocarpus.

```{r warning=FALSE, message=FALSE}
compute_perm <- function(og_sc_sig, species_select, n_perm = 1000){
        df_rand <- 
                og_sc_sig |>
                rsample::permutations(permute = species_select,
                                      times = n_perm,
                                      apparent = TRUE)

        count_same_fast_select <- function(df_rand, species_select) {
                df_rand_select <- df_rand |> dplyr::select(Ec, !!sym(species_select))
                same <- apply(df_rand_select, 1, function(x) length(unique(x)) == 1)
                sum(same)
        }
        df_out <-
                df_rand |>
                dplyr::mutate(n_same = purrr::map_int(splits, ~ count_same_fast_select(analysis(.x), species_select = species_select))) |>
                dplyr::select(id, n_same) |> 
                dplyr::mutate(data = dplyr::case_when(
                        id == "Apparent" ~ "empirical",
                        .default = "permutation"
                        ))
        df_out <- split(df_out, df_out$data)
        return(df_out)
}

Ec_Spro_perm <- compute_perm(og_sc_sig, species_select = "Spro")
Ec_Und_perm <- compute_perm(og_sc_sig, species_select = "Und")
Ec_Dher_perm <- compute_perm(og_sc_sig, species_select = "Dher")
Ec_Ddud_perm <- compute_perm(og_sc_sig, species_select = "Ddud")
Ec_Si_perm <- compute_perm(og_sc_sig, species_select = "Si")
```

```{r warning=FALSE, message=FALSE}
Ec_vall_perm <-
        Ec_Spro_perm$permutation |> dplyr::mutate(sp = "Spro") |>
        rbind(
                Ec_Und_perm$permutation |> dplyr::mutate(sp = "Und")
        ) |>
        rbind(
                Ec_Dher_perm$permutation |> dplyr::mutate(sp = "Dher")
        ) |>
        rbind(
                Ec_Ddud_perm$permutation |> dplyr::mutate(sp = "Ddud")
        ) |>
        rbind(
                Ec_Si_perm$permutation |> dplyr::mutate(sp = "Si")
        ) |>
        dplyr::mutate(sp = factor(sp, levels = c("Spro", "Und", "Ddud", "Dher", "Si")))

Ec_vall_emp <-
        Ec_Spro_perm$empirical |> dplyr::mutate(sp = "Spro") |>
        rbind(
                Ec_Und_perm$empirical |> dplyr::mutate(sp = "Und")
        ) |>
        rbind(
                Ec_Dher_perm$empirical |> dplyr::mutate(sp = "Dher")
        ) |>
        rbind(
                Ec_Ddud_perm$empirical |> dplyr::mutate(sp = "Ddud")
        ) |>
        rbind(
                Ec_Si_perm$empirical |> dplyr::mutate(sp = "Si")
        ) |>
        dplyr::mutate(sp = factor(sp, levels = c("Spro", "Und", "Ddud", "Dher", "Si")))
```

```{r warning=FALSE, message=FALSE, fig.width=3, fig.height=2.5}
Ec_vall_perm |>
        ggplot2::ggplot(aes(x = sp, y = n_same)) +
        see::geom_violinhalf(colour = "black", fill = "lavender", width = 1.5) +
        ggplot2::geom_point(data = Ec_vall_emp, colour = "red", shape = 8) +
        ggplot2::theme_classic() +
        ggplot2::scale_x_discrete(limits=rev) +
        theme(panel.grid.major.y = element_line(colour = "grey90")) +
        ggplot2::coord_flip() +
        ggplot2::labs(x = "species", y = "no. orthologues w/ conserved signature")
```


Check for normality

```{r warning=FALSE, message=FALSE, fig.width=5, fig.height=4}
Ec_vall_perm |> 
        ggpubr::ggqqplot(x = "n_same") +
        ggplot2::facet_wrap(.~sp, scales = "free_y")

Ec_vall_perm |>
        dplyr::summarise(
                shapiro = list(shapiro.test(n_same)),
                .by = "sp"
                ) |>
        dplyr::mutate(
                statistic = purrr::map_dbl(shapiro, ~ .x$statistic),
                p_value = purrr::map_dbl(shapiro, ~ .x$p.value)
                )

```

Since it is normally distributed, we can compute the p-values

```{r warning=FALSE, message=FALSE}
# computing from the permutations
Ec_vall_perm |>
        dplyr::inner_join(Ec_vall_emp, by = "sp", suffix = c("_perm", "_emp")) |>
        dplyr::summarise(
                p_value = mean(n_same_perm >= n_same_emp),
                .by = sp
                )
```

```{r warning=FALSE, message=FALSE}
# since the permutations approximates the normal distrbution, let's use this as well
# summarise mean and sd from permutations
# this is the Z-test
perm_stats <- 
        Ec_vall_perm |>
        dplyr::summarise(
                mean_perm = mean(n_same),
                sd_perm = sd(n_same),
                .by = sp
                )

emp_pvals <-
        Ec_vall_emp |>
        dplyr::left_join(perm_stats, by = "sp") |>
        dplyr::mutate(
                z = (n_same - mean_perm) / sd_perm,
                p_value = 1 - pnorm(z)  # right-tailed
                ) |>
        dplyr::select(sp, empirical = n_same, mean_perm, sd_perm, z, p_value)

emp_pvals
```

With 1000 permutations

```R
> emp_pvals
# A tibble: 5 × 6
  sp    empirical mean_perm sd_perm     z   p_value
  <fct>     <int>     <dbl>   <dbl> <dbl>     <dbl>
1 Spro       1847     1716.    13.5  9.68 0        
2 Und        1351     1195.    15.6 10.0  0        
3 Dher       1588     1430.    14.9 10.6  0        
4 Ddud       1785     1600.    15.0 12.3  0        
5 Si         1123     1058.    15.6  4.19 0.0000141
```

Since the permutations are normally distributed, we can assume the mean and sd of the sample. With this, we can compute the p-value using the z-test. The z-test is used rather than the t-test because n is large and the central limit theorem should apply well.

BUT it is not standard so I will just use the permutation p-value.

### Permutations across all species (same signature for all)

If I resample all

```{r warning=FALSE, message=FALSE}
n_perm = 100
perm_list <- 1:n_perm %>%
  purrr::map(~ og_sc_sig %>%
        dplyr::mutate(across(c(Ddud, Dher, Ec, Und, Si, Spro), ~ sample(.))) )

# create a rsample-like object to apply the count function
og_sc_sig.rand.all <- tibble(
  splits = perm_list,
  id = paste0("Permutation", 1:n_perm)
)
```


```{r}
count_same_fast <- function(df) {
  same <- apply(df |> dplyr::select(Ddud:Spro), 1, function(x) length(unique(x)) == 1)
  sum(same)
}

og_sc_sig.rand.res.all <- og_sc_sig.rand.all |>
        dplyr::mutate(n_same = map_int(splits, ~ count_same_fast(.x))) |>
        dplyr::select(id, n_same) |>
        # add empirical data
        rbind(
                tibble::tibble(
                        id = "Apparent",
                        n_same = 264
                )
        ) |> 
        dplyr::mutate(data = dplyr::case_when(
                id == "Apparent" ~ "empirical",
                .default = "permutation"
        )) 
```

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
og_sc_sig.rand.res.all |>
        dplyr::filter(data == "permutation") |>
        ggplot2::ggplot(aes(n_same, fill = data)) +
        ggplot2::geom_histogram(colour = "black", fill = "lavender") +
        ggplot2::geom_vline(xintercept = 264, colour = "red", linetype = 2) +
        ggplot2::theme_classic() +
        ggplot2::labs(
                x = "orthologue w/ conserved signature",
                title = "Resampling all species"
        )
```

Let's try this with 1000 permutations.

```{r warning=FALSE, message=FALSE}
n_perm = 1000
perm_list <- 1:n_perm %>%
  purrr::map(~ og_sc_sig %>%
        dplyr::mutate(across(c(Ddud, Dher, Ec, Und, Si, Spro), ~ sample(.))) )

# create a rsample-like object to apply the count function
og_sc_sig.rand.all.1k <- tibble(
  splits = perm_list,
  id = paste0("Permutation", 1:n_perm)
)
```


```{r}
count_same_fast <- function(df) {
  same <- apply(df |> dplyr::select(Ddud:Spro), 1, function(x) length(unique(x)) == 1)
  sum(same)
}

og_sc_sig.rand.res.all.1k <- og_sc_sig.rand.all.1k |>
        dplyr::mutate(n_same = map_int(splits, ~ count_same_fast(.x))) |>
        dplyr::select(id, n_same) |>
        # add empirical data
        rbind(
                tibble::tibble(
                        id = "Apparent",
                        n_same = 264
                )
        ) |> 
        dplyr::mutate(data = dplyr::case_when(
                id == "Apparent" ~ "empirical",
                .default = "permutation"
        )) 
```

```{r warning=FALSE, message=FALSE, fig.width=4, fig.height=2.5}
og_sc_sig.rand.res.all.1k |> 
        ggplot2::ggplot(aes(n_same, fill = data)) +
        ggplot2::geom_histogram(colour = "black") +
        ggrepel::geom_label_repel(
                data = tibble(x = 264, label = "empirical = 264"),
                ggplot2::aes(x = x, y = 1,
                             label = label),
                stat = "unique",
                size = 3,
                nudge_y = 30,
                inherit.aes = FALSE) +
        ggplot2::theme_classic() +
        ggplot2::labs(
                x = "orthologue w/ conserved signature",
                title = "Resampling all species",
                y = "count"
        )
```

```{r}
og_sc_sig.rand.res.all.1k.perm <- 
        og_sc_sig.rand.res.all.1k |>
        dplyr::filter(data == "permutation")
```

Check for normality

```{r warning=FALSE, message=FALSE, fig.width=3, fig.height=3}
og_sc_sig.rand.res.all.1k.perm |> 
        ggpubr::ggqqplot(x = "n_same")
# shapiro.test(og_sc_sig.rand.res.all.1k.perm$n_same)
```

Since it is roughly normally distributed, we can compute the p-values

```{r warning=FALSE, message=FALSE}
# computing from the permutations
og_sc_sig.rand.res.all.1k.perm |>
        dplyr::summarise(p_value = mean(n_same >= 264))
```

```{r warning=FALSE, message=FALSE}
# since the permutations approximates the normal distrbution, let's use this as well
# summarise mean and sd from permutations
# this is the Z-test
perm_stats <- 
        og_sc_sig.rand.res.all.1k.perm |>
        dplyr::summarise(
                mean_perm = mean(n_same),
                sd_perm = sd(n_same)
                )

emp_pvals <-
        perm_stats |>
        dplyr::mutate(
                z = (264 - mean_perm) / sd_perm,
                p_value = 1 - pnorm(z)  # right-tailed
                )

emp_pvals
```

With 10000 permutations

```R
> emp_pvals
# A tibble: 1 × 4
  mean_perm sd_perm     z p_value
      <dbl>   <dbl> <dbl>   <dbl>
1      122.    9.98  14.2       0
```

For 10k permutations, both Ec_vall_perm and og_sc_sig.rand.res.all have been saved as `.rds` files in data.

```{r}
Ec_vall_perm <- readr::read_rds("data/Ec_vall_perm.rds")
og_sc_sig.rand.res.all <- readr::read_rds("data/og_sc_perm.rds")
```
