---
title: "Signature conservation across species between sex chromosomes and autosomes"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

Signatures are more conserved between species than permutations. The next question is how this looks if we analyse genes in the sex chromosomes separately from autosomes.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(tidymodels) # for rsample
library(readxl)
```

Read and make tx2gene file to replace gene IDs for some species

```{r warning=FALSE, message=FALSE}
Und_tx2gene <- readr::read_tsv("data/TPM/Und_tx2gene.tsv",col_select = c(1,2), col_names = c("tx", "gene_id"))
Si_tx2gene <- readr::read_tsv("data/TPM/Schizocladia-ischiensis.tsv",col_select = c(1,2))
```

Read signature master file

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")

Ec_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Ddud_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Dher_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))
Spro_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("Signature"))

# species where renaming is needed

Und_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        # rename gene_id
        dplyr::left_join(Und_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
Si_sig <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        # rename gene_id
        dplyr::left_join(Si_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("Signature"))
```

Read and process orthogroup data to obtain one-to-one orthologues

```{r warning=FALSE, message=FALSE}
# filter longest isoform in Ectocarpus
library(rtracklayer)

Ec_data.gtf <- rtracklayer::import("data/annotation/Ec32_V5.longest_mRNA.SDR_F.AGAT.mod.gtf")
Ec_data.gtf.df <- Ec_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id) |>
        dplyr::distinct()
```

```{r warning=FALSE, message=FALSE}
og_data <-
        readr::read_tsv("data/Orthogroups.tsv") |>
        tidyr::drop_na() |>
        tidyr::pivot_longer(
                cols = -Orthogroup,
                names_to = "species",
                values_to = "gene_id"
                ) |>
        dplyr::filter(!is.na(gene_id)) |>
        tidyr::separate_rows(gene_id, sep = ",\\s*") |>
        dplyr::filter(gene_id != "") |>
        # filter isoforms
        dplyr::filter(!stringr::str_detect(gene_id, "\\.t[0-9]+$") | stringr::str_detect(gene_id, "\\.t1$")) |>
        dplyr::filter(
                (species == "PUBLIC_Ectocarpus-sp7_proteins" & gene_id %in% Ec_data.gtf.df$gene_id) |
                        (species != "PUBLIC_Ectocarpus-sp7_proteins")
        )

# get single copy data
og_sc.pre <- 
        og_data |>
        dplyr::summarise(n = dplyr::n(), .by = c(Orthogroup, species)) |>
        dplyr::filter(sum(n) == 6, .by = c(Orthogroup)) |>
        dplyr::distinct(Orthogroup)
og_sc <-
        og_data |>
        dplyr::filter(Orthogroup %in% og_sc.pre$Orthogroup)
og_sc.edit <-
        og_sc |>
        dplyr::mutate(
                gene_id = stringr::str_remove(gene_id, "\\.t1$"),
                gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_"),
                species = dplyr::case_when(
                        stringr::str_detect(species,"dudresnayi") ~ "Ddud",
                        stringr::str_detect(species,"herbacea") ~ "Dher",
                        stringr::str_detect(species,"Ectocarpus") ~ "Ec",
                        stringr::str_detect(species,"Undaria") ~ "Und",
                        stringr::str_detect(species,"Scytosiphon") ~ "Spro",
                        stringr::str_detect(species,"Schizocladia") ~ "Si"
                ))
```


Where are the sex chromosomes? According to Barrera-Redondo et al.

```
Ec              chr_13
Spro            chr_13
Und             chr_23
Dd              chr_02
Dher            chr_03
Ddud_pseudo     chr_03
```

Let's begin with Ectocarpus

```{r warning=FALSE, message=FALSE}
library(rtracklayer)

Ec_data.gtf <- rtracklayer::import("data/annotation/Ec32_V5.longest_mRNA.SDR_F.AGAT.mod.gtf")
Ec_data.gtf.df <- Ec_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id) |>
        dplyr::distinct()

og_sc_chr_id <- 
        og_sc |> 
        dplyr::filter(species == "PUBLIC_Ectocarpus-sp7_proteins") |>
        dplyr::left_join(Ec_data.gtf.df) |>
        tidyr::drop_na() |>
        dplyr::select(Orthogroup, chr = seqnames)
```

Combine orthologue and signature data

```{r warning=FALSE, message=FALSE}
sig.all <- 
        Ec_sig |> dplyr::select(1,Signature = Signature.Female) |>
        rbind(Ddud_sig) |>
        rbind(Dher_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Spro_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Und_sig |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Si_sig)

og_sc_sig <- 
        og_sc.edit |>
        dplyr::left_join(sig.all) |>
        dplyr::distinct(Orthogroup, species, .keep_all = TRUE) |>  # keep first match only
        tidyr::pivot_wider(!gene_id, names_from = species, values_from = Signature) |>
        tidyr::drop_na()
```


```{r warning=FALSE, message=FALSE}
og_sc_sig_chr <- og_sc_sig |> dplyr::left_join(og_sc_chr_id)
```


```{r warning=FALSE, message=FALSE}
count_same_by_chr <- function(df_rand, species, species_select) {
        df_rand |>
                dplyr::select(chr, !!sym(species), !!sym(species_select)) |>
                # group_by(chr) |>
                dplyr::summarise(same_count = sum(!!sym(species) == !!sym(species_select)),
                          total_count = dplyr::n(), 
                          same_perc = 100*same_count/total_count, .by = chr) |>
                dplyr::mutate(sp = species_select)
        }
```

```{r warning=FALSE, message=FALSE}
og_sc_sig_same_by_chr <-
        count_same_by_chr(og_sc_sig_chr, "Ec", "Ddud") |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ec", "Dher")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ec", "Und")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ec", "Si")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ec", "Spro")) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Spro", "Und", "Ddud", "Dher", "Si"
        )))
```

As you can see, the sex chromosome (chr_13) has a different profile from the autosomes.

```{r warning=FALSE, message=FALSE, fig.width=3, fig.height=4}
og_sc_sig_same_by_chr |>
        dplyr::filter(chr != "HiC_scaffold_28") |> # drop scaffold
        ggplot2::ggplot(aes(x = chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = chr, y = same_perc, group = chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_line(alpha = 0.8, linetype = 2) +
        ggplot2::coord_flip() +
        ggplot2::theme_classic() +
        ggplot2::labs(x = "Ectocarpus chromosomes", y = "signatures conserved (%)", colour = "species") +
        ggsci::scale_color_npg()
```

```{r warning=FALSE, message=FALSE, fig.width=5, fig.height=2.5}
og_sc_sig_same_by_chr |>
        dplyr::filter(chr != "HiC_scaffold_28") |> # drop scaffold
        ggplot2::ggplot(aes(x = chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = chr, y = same_perc, group = chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Ectocarpus chromosomes", y = "signatures conserved (%)", colour = "species") +
        ggsci::scale_color_npg()
```

### Summarising by autosome vs sex chromosomes.

It should be noted that while we split the genes in the focal species to sex or autosome, the orthologues in the compared species can belong to any chromosome type. This is because there is a fast turnover of genes in the sex chromosomes so there are very few genes that are (1) even detected within the one-to-one ortholog set that also (2) remain in the sex chromosomes.

```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=3}
og_sc_sig_same_by_chr_autosome_summary <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "chr_13" ~ "sex (chr_13)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))

og_sc_sig_same_by_chr_autosome_summary |>
        ggplot2::ggplot(aes(x = sex_chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = sex_chr, y = same_perc, group = sex_chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_point(aes(x = sex_chr, y = same_perc, group = sex_chr)) +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        # ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Ec chromosome", y = "signatures conserved (%)", colour = "species") +
        ggsci::scale_color_npg() +
        ggplot2::labs(title = "Ectocarpus sp.7 chr_13")

og_sc_sig_same_by_chr_autosome_summary.Ec <- og_sc_sig_same_by_chr_autosome_summary
```

```{r}
test_input <-
        og_sc_sig_same_by_chr_autosome_summary |>
        tidyr::pivot_wider(names_from = sex_chr, values_from = same_perc)

wilcox.test(test_input$autosome,
            test_input$`sex (chr_13)`, 
            paired = TRUE,
            alternative = "greater")

t.test(test_input$autosome,
            test_input$`sex (chr_13)`, 
            paired = TRUE,
            alternative = "greater")
```

```{r warning=FALSE, message=FALSE}
# use consistent colours for other species
ggsci::pal_npg() |> scales::show_col()
ggsci::pal_npg()(6)[1:6]

color_table <-
        tibble::tibble(
                sp = c("Spro", "Und", "Ddud", "Dher", "Si", "Ec"),
                Colour = ggsci::pal_npg()(6)[1:6]
        ) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Spro", "Und", "Ddud", "Dher", "Si", "Ec"
        )))
```

```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=3}
# Spro chr_13
Spro_data.gtf <- rtracklayer::import("data/annotation/convert_gffread/Scytosiphon-promiscuus_MALE.chr.gffread.gff")
Spro_data.gtf.df <- Spro_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = ID) |>
        dplyr::distinct()

og_sc_chr_id <- 
        og_sc |> 
        dplyr::filter(species == "Scytosiphon-promiscuus_MALE_proteins") |>
        dplyr::mutate(gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_")) |>
        dplyr::left_join(Spro_data.gtf.df) |>
        tidyr::drop_na() |>
        dplyr::select(Orthogroup, chr = seqnames)

og_sc_sig_chr <- og_sc_sig |> dplyr::left_join(og_sc_chr_id)

og_sc_sig_same_by_chr <-
        count_same_by_chr(og_sc_sig_chr, "Spro", "Ddud") |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Spro", "Dher")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Spro", "Und")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Spro", "Si")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Spro", "Ec")) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Und", "Ddud", "Dher", "Si"
        )))


og_sc_sig_same_by_chr_autosome_summary <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "chr_13" ~ "sex (chr_13)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))

og_sc_sig_same_by_chr_autosome_summary |>
        ggplot2::ggplot(aes(x = sex_chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = sex_chr, y = same_perc, group = sex_chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_point(aes(x = sex_chr, y = same_perc, group = sex_chr)) +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        # ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Spro chromosome", y = "signatures conserved (%)", colour = "species") +
        ggplot2::scale_colour_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "Scytosiphon promiscuus chr_13")
# Wilcoxon signed rank exact test, V = 12, p-value = 0.1562

og_sc_sig_same_by_chr_autosome_summary.Spro <- og_sc_sig_same_by_chr_autosome_summary
```

```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=3}
# Und HiC_scaffold_23
# there are apparently 30 'chromosome-level' scaffolds.
# I confirmed this with seqkit.
Und_data.gtf <- rtracklayer::import("data/annotation/PUBLIC_Undaria-pinnatifida_MALE.jv3.gtf")
Und_data.gtf.df <- Und_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = transcript_id) |>
        dplyr::distinct()

og_sc_chr_id <- 
        og_sc |> 
        dplyr::filter(species == "PUBLIC_Undaria-pinnatifida_MALE.pep") |>
        dplyr::mutate(gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_")) |>
        dplyr::left_join(Und_data.gtf.df) |>
        tidyr::drop_na() |>
        dplyr::select(Orthogroup, chr = seqnames)

og_sc_sig_chr <- og_sc_sig |> dplyr::left_join(og_sc_chr_id)

og_sc_sig_same_by_chr <-
        count_same_by_chr(og_sc_sig_chr, "Und", "Spro") |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Und", "Ddud")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Und", "Dher")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Und", "Si")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Und", "Ec")) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Spro", "Ddud", "Dher", "Si"
        )))


og_sc_sig_same_by_chr_autosome_summary <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "HiC_scaffold_23" ~ "sex (chr_23)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))

og_sc_sig_same_by_chr_autosome_summary |>
        ggplot2::ggplot(aes(x = sex_chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = sex_chr, y = same_perc, group = sex_chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_point(aes(x = sex_chr, y = same_perc, group = sex_chr)) +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        # ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Und chromosome", y = "signatures conserved (%)", colour = "species") +
        ggplot2::scale_colour_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "Undaria pinnatifida chr_23 (HiC scaffold)")
# Wilcoxon signed rank exact test, V = 13, p-value = 0.09375

og_sc_sig_same_by_chr_autosome_summary.Und <- og_sc_sig_same_by_chr_autosome_summary
```


```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=3}
# Dher chr_03
Dher_data.gtf <- rtracklayer::import("data/annotation/convert_gffread/Desmarestia-herbacea.reference.gt.AGAT.gffread.gff")
Dher_data.gtf.df <- Dher_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = ID) |>
        dplyr::distinct()

og_sc_chr_id <- 
        og_sc |> 
        dplyr::filter(species == "Desmarestia-herbacea_MALE_proteins") |>
        dplyr::mutate(gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_")) |>
        dplyr::left_join(Dher_data.gtf.df) |>
        tidyr::drop_na() |>
        dplyr::select(Orthogroup, chr = seqnames)

og_sc_sig_chr <- og_sc_sig |> dplyr::left_join(og_sc_chr_id)

og_sc_sig_same_by_chr <-
        count_same_by_chr(og_sc_sig_chr, "Dher", "Ddud") |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Dher", "Spro")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Dher", "Und")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Dher", "Si")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Dher", "Ec")) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Und", "Ddud", "Spro", "Si"
        )))


og_sc_sig_same_by_chr_autosome_summary <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "chr_03" ~ "sex (chr_03)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))

og_sc_sig_same_by_chr_autosome_summary |>
        ggplot2::ggplot(aes(x = sex_chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = sex_chr, y = same_perc, group = sex_chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_point(aes(x = sex_chr, y = same_perc, group = sex_chr)) +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        # ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Dher chromosome", y = "signatures conserved (%)", colour = "species") +
        ggplot2::scale_colour_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "Desmarestia herbacea chr_03")
# Wilcoxon signed rank exact test, V = 15, p-value = 0.03125

og_sc_sig_same_by_chr_autosome_summary.Dher <- og_sc_sig_same_by_chr_autosome_summary
```

### And in the former sex chromosome of D. dudresnayi

```{r warning=FALSE, message=FALSE, fig.width=3.5, fig.height=3}
# Ddud pseudo-chr_03
Ddud_data.gtf <- rtracklayer::import("data/annotation/Desmarestia-dudresnayi.cleaned2025.jv.gffread.gtf")
Ddud_data.gtf.df <- Ddud_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id = transcript_id) |>
        dplyr::distinct()

og_sc_chr_id <- 
        og_sc |> 
        dplyr::filter(species == "Desmarestia-dudresnayi_BRAKER_proteins") |>
        dplyr::left_join(Ddud_data.gtf.df) |>
        tidyr::drop_na() |>
        dplyr::select(Orthogroup, chr = seqnames)

og_sc_sig_chr <- og_sc_sig |> dplyr::left_join(og_sc_chr_id)

og_sc_sig_same_by_chr <-
        count_same_by_chr(og_sc_sig_chr, "Ddud", "Dher") |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ddud", "Spro")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ddud", "Und")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ddud", "Si")) |>
        rbind(count_same_by_chr(og_sc_sig_chr, "Ddud", "Ec")) |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Und", "Dher", "Spro", "Si"
        )))


og_sc_sig_same_by_chr_autosome_summary <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "chr_03" ~ "sex (chr_03)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))

og_sc_sig_same_by_chr_autosome_summary |>
        # dplyr::filter(sp != "Si") |>
        ggplot2::ggplot(aes(x = sex_chr, y = same_perc, colour = sp, group = sp)) +
        ggplot2::geom_boxplot(aes(x = sex_chr, y = same_perc, group = sex_chr),
                              # outliers = FALSE,
                              show.legend = FALSE,
                              fill = "lavender") +
        ggplot2::geom_point(aes(x = sex_chr, y = same_perc, group = sex_chr)) +
        ggplot2::geom_line(alpha = 1, linetype = 2) +
        ggplot2::theme_classic() +
        # ggplot2::theme(axis.text.x = element_text(angle = 90)) +
        ggplot2::labs(x = "Ddud chromosome", y = "signatures conserved (%)", colour = "species") +
        ggplot2::scale_colour_manual(values = setNames(color_table$Colour, color_table$sp)) +
        ggplot2::labs(title = "Desmarestia dudresnayi chr_03 (pseudo)")

# Wilcoxon signed rank exact test, V = 10, p-value = 0.3125


og_sc_sig_same_by_chr_autosome_summary.Ddud <- 
        og_sc_sig_same_by_chr |>
        dplyr::mutate(sex_chr = dplyr::case_match(
                chr,
                "chr_03" ~ "pseudosex (chr_03)",
                .default = "autosome"
        )) |>
        dplyr::summarise(same_perc = 100*sum(same_count)/sum(total_count),
                         .by = c(sp, sex_chr))
```

### Plotting all species in the same plot.

```{r warning=FALSE, message=FALSE, fig.width=5, fig.height=4}
og_sc_sig_same_by_chr_autosome_summary.combined <-
        rbind(
                og_sc_sig_same_by_chr_autosome_summary.Ec |>
                        dplyr::mutate(species = "Ec") |>
                        dplyr::mutate(sex_type = "dioicous"),
                og_sc_sig_same_by_chr_autosome_summary.Spro |>
                        dplyr::mutate(species = "Spro") |>
                        dplyr::mutate(sex_type = "dioicous")
        ) |>
        rbind(
                og_sc_sig_same_by_chr_autosome_summary.Und |>
                        dplyr::mutate(species = "Und") |>
                        dplyr::mutate(sex_type = "dioicous")
        ) |>
        rbind(
                og_sc_sig_same_by_chr_autosome_summary.Dher |>
                        dplyr::mutate(species = "Dher") |>
                        dplyr::mutate(sex_type = "dioicous")
        ) |>
        rbind(
                og_sc_sig_same_by_chr_autosome_summary.Ddud |>
                        dplyr::mutate(species = "Ddud") |>
                        dplyr::mutate(sex_type = "monoicous")
        ) |>
        dplyr::mutate(species = factor(species, levels = c(
                "Ec", "Spro", "Und", "Dher", "Ddud"
        )))

og_sc_sig_same_by_chr_autosome_summary.combined.fill <- og_sc_sig_same_by_chr_autosome_summary.combined |> dplyr::distinct(species, sex_type)

# dataframe for shading background only for monoicous
rect_df <- og_sc_sig_same_by_chr_autosome_summary.combined.fill |>
        dplyr::filter(sex_type == "monoicous") |>
        dplyr::mutate(xmin = -Inf, xmax = Inf, ymin = -Inf, ymax = Inf) |>
        dplyr::mutate(species = factor(species, levels = c(
                "Ec", "Spro", "Und", "Dher", "Ddud"
        )))

og_sc_sig_same_by_chr_autosome_summary.combined |>
  ggplot2::ggplot(aes(
    x = sex_chr, y = same_perc, colour = sp, group = sp)) +
  # shading for monoicous species only
  ggplot2::geom_rect(
    data = rect_df,
    aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE,
    fill = "grey", alpha = 0.2
  ) +
  # the rest
  ggplot2::geom_boxplot(aes(group = sex_chr),
                        show.legend = FALSE,
                        fill = "lavender") +
  ggplot2::geom_point() +
  ggplot2::geom_line(alpha = 1, linetype = 2) +
  ggplot2::theme_classic() +
  ggplot2::scale_colour_manual(values = setNames(color_table$Colour, color_table$sp)) +
  ggplot2::labs(title = "all species") +
  ggplot2::facet_grid(. ~ species, scales = "free_x") +
  ggplot2::labs(x = "chromosome", y = "same signatures (%)", colour = "species") +
  ggplot2::guides(x = guide_axis(angle = 90))
```


```{r warning=FALSE, message=FALSE, fig.width=6, fig.height=2.5}
og_sc_sig_same_by_chr_autosome_summary.combined |>
        dplyr::mutate(sp = factor(sp, levels = c(
                "Ec", "Spro", "Und", "Dher", "Ddud", "Si"
        ))) |>
        dplyr::mutate(chr_type = dplyr::case_when(
                sex_chr == "autosome" ~ "autosome",
                .default = "sex chromosome"
        )) |>
        ggplot2::ggplot(
                aes(x = sp, y = same_perc, fill = chr_type, group = sex_chr)
        ) +
        
        ggplot2::geom_rect(
                data = rect_df,
                aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
                inherit.aes = FALSE,
                fill = "grey", alpha = 0.3
        ) +
        
        ggplot2::geom_col(position = "dodge", colour = "black", width = 0.8) +
        
        ggplot2::theme_classic() +
        ggplot2::scale_fill_manual(
                values = c("autosome" = "steelblue", "sex chromosome" = "white")) +
        ggplot2::labs(
                title = "all species",
                x = "species compared",
                y = "same signatures (%)",
                fill = "orthologs from"
        ) +
        ggplot2::facet_grid(. ~ species, scales = "free_x", ) +
        ggplot2::guides(x = ggplot2::guide_axis(angle = 90))

```