---
title: "profile of one-to-one orthologues"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

In the signature_conservation document, we used one-to-one orthologues from orthofinder. We were wondering, however, if these genes share a different profile to the rest of the genes.

```{r warning=FALSE, message=FALSE}
library(tidyverse)
library(ggrastr)
library(readxl)
library(scales)
```

Read and process orthogroup data to obtain one-to-one orthologues

Get a single isoform per gene (post-hoc).

```{r warning=FALSE, message=FALSE}
# filter longest isoform in Ectocarpus
library(rtracklayer)

Ec_data.gtf <- rtracklayer::import("data/annotation/Ec32_V5.longest_mRNA.SDR_F.AGAT.mod.gtf")
Ec_data.gtf.df <- Ec_data.gtf |>
        rtracklayer::as.data.frame() |>
        dplyr::select(seqnames, gene_id) |>
        dplyr::distinct()
```

```{r warning=FALSE, message=FALSE}
og_data <-
        readr::read_tsv("data/Orthogroups.tsv") |>
        tidyr::drop_na() |>
        tidyr::pivot_longer(
                cols = -Orthogroup,
                names_to = "species",
                values_to = "gene_id"
                ) |>
        dplyr::filter(!is.na(gene_id)) |>
        tidyr::separate_rows(gene_id, sep = ",\\s*") |>
        dplyr::filter(gene_id != "") |>
        # filter isoforms
        dplyr::filter(!stringr::str_detect(gene_id, "\\.t[0-9]+$") | stringr::str_detect(gene_id, "\\.t1$")) |>
        dplyr::filter(
                (species == "PUBLIC_Ectocarpus-sp7_proteins" & gene_id %in% Ec_data.gtf.df$gene_id) |
                        (species != "PUBLIC_Ectocarpus-sp7_proteins")
        )

# get single copy data
og_sc.pre <- 
        og_data |>
        dplyr::summarise(n = dplyr::n(), .by = c(Orthogroup, species)) |>
        dplyr::filter(sum(n) == 6, .by = c(Orthogroup)) |>
        dplyr::distinct(Orthogroup)
og_sc <-
        og_data |>
        dplyr::filter(Orthogroup %in% og_sc.pre$Orthogroup)
og_sc.edit <-
        og_sc |>
        dplyr::mutate(
                gene_id = stringr::str_remove(gene_id, "\\.t1$"),
                gene_id = stringr::str_replace(gene_id, "prot_", "mRNA_"),
                species = dplyr::case_when(
                        stringr::str_detect(species,"dudresnayi") ~ "Ddud",
                        stringr::str_detect(species,"herbacea") ~ "Dher",
                        stringr::str_detect(species,"Ectocarpus") ~ "Ec",
                        stringr::str_detect(species,"Undaria") ~ "Und",
                        stringr::str_detect(species,"Scytosiphon") ~ "Spro",
                        stringr::str_detect(species,"Schizocladia") ~ "Si"
                ))
```

Obtain gene age and signature information

```{r warning=FALSE, message=FALSE}
readxl::excel_sheets("data/dataset_S3_master_tables.xlsx")
Ec_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Ectocarpus") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Ddud_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-dudresnayi") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Dher_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Desmarestia-herbacea") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
Spro_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Scitosyphon") |>
        dplyr::select(gene_id, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
```

```{r warning=FALSE, message=FALSE}
# species where renaming is needed
Und_tx2gene <- readr::read_tsv("data/TPM/Und_tx2gene.tsv",col_select = c(1,2), col_names = c("tx", "gene_id"))
Si_tx2gene <- readr::read_tsv("data/TPM/Schizocladia-ischiensis.tsv",col_select = c(1,2))


Und_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Undaria") |>
        dplyr::left_join(Und_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)

Si_ps <- readxl::read_xlsx("data/dataset_S3_master_tables.xlsx", sheet = "Schizocladia") |>
        dplyr::left_join(Si_tx2gene) |> dplyr::select(-gene_id) |>
        dplyr::select(gene_id = tx, dplyr::contains("TPM"), dplyr::contains("Signature"), rank)
```

Combine orthologue and signature data

```{r warning=FALSE, message=FALSE}
sig.all <- 
        Ec_ps |> dplyr::select(1,Signature = Signature.Female) |>
        rbind(Ddud_ps |> dplyr::select(1,Signature)) |>
        rbind(Dher_ps |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Spro_ps |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Und_ps |> dplyr::select(1,Signature = Signature.Female)) |>
        rbind(Si_ps |> dplyr::select(1,Signature))

og_sc_sig <- 
        og_sc.edit |>
        dplyr::left_join(sig.all) |>
        dplyr::distinct(Orthogroup, species, .keep_all = TRUE) |>  # keep first match only
        tidyr::pivot_wider(!gene_id, names_from = species, values_from = Signature) |>
        tidyr::drop_na()
```

```{r warning=FALSE, message=FALSE}
og_sc_sig.6 <- 
        og_sc_sig |> 
        dplyr::mutate(all_same = n_distinct(c_across(Ddud:Spro)) == 1,
                      .by = Orthogroup)

og_sc_sig.6.conserved <-
        og_sc_sig.6 |> dplyr::filter(all_same == TRUE)


og_sc.edit.strict <- 
        og_sc.edit |> dplyr::filter(Orthogroup %in% og_sc_sig.6$Orthogroup)
og_sc.edit.strict.conserved <- 
        og_sc.edit |> dplyr::filter(Orthogroup %in% og_sc_sig.6.conserved$Orthogroup)
```

```{r}
# without accounting for all genes that are labelled by signature and age across species
nrow(og_sc.edit)/6
# taking this into account
nrow(og_sc.edit.strict)/6
```

## Gene age

Begin with Ec

```{r warning=FALSE, message=FALSE, fig.width=2, fig.height=2.5}
Ec_ps.processed <-
        Ec_ps |>
        tidyr::drop_na() |>
        dplyr::mutate(rank = as.numeric(rank)) |>
        tidyr::drop_na(rank) |>
        dplyr::mutate(single_copy_og = dplyr::case_when(
                gene_id %in% og_sc.edit.strict$gene_id ~ "yes",
                .default = "no"
        )) 
Ec_ps.processed |>
        ggplot2::ggplot(aes(x = single_copy_og, y = rank, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "1-to-1 orthologue", y = "phylorank") +
        ggplot2::scale_color_viridis_d()
Ec_ps.processed |>
        ggplot2::ggplot(aes(x = single_copy_og, y = rank, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1),dpi = 300) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "1-to-1 orthologue", y = "phylorank") +
        ggplot2::scale_color_viridis_d() +
        # ggpubr::stat_kruskal_test(label = "p.adj", size = 2.5) +
        EnvStats::stat_n_text(size = 2.5)
```

```{r warning=FALSE, message=FALSE, fig.width=2, fig.height=2.5}
Ec_ps.processed2 <-
        Ec_ps |>
        tidyr::drop_na() |>
        dplyr::mutate(rank = as.numeric(rank)) |>
        tidyr::drop_na(rank) |>
        dplyr::mutate(single_copy_og = dplyr::case_when(
                gene_id %in% og_sc.edit.strict.conserved$gene_id ~ "yes",
                .default = "no"
        )) 
Ec_ps.processed2 |>
        ggplot2::ggplot(aes(x = single_copy_og, y = rank, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "conserved 1-to-1 orthologue", y = "phylorank") +
        ggplot2::scale_color_viridis_d()
Ec_ps.processed2 |>
        ggplot2::ggplot(aes(x = single_copy_og, y = rank, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1),dpi = 300) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "conserved 1-to-1 orthologue", y = "phylorank") +
        ggplot2::scale_color_viridis_d() +
        ggpubr::stat_kruskal_test(label = "p.adj", size = 2.5) +
        EnvStats::stat_n_text(size = 2.5)
```

## Tau

using Ectocarpus data from hourglass

```{r warning=FALSE, message=FALSE}
library(tidyr); library(duckplyr)
options(duckdb.materialize_message = FALSE)
```


```{r warning=FALSE, message=FALSE}
# Ec_TPM <- readr::read_tsv("../atlas_ref/data/expression_data_processed/Ec_TPM.tsv") |>
#         tibble::column_to_rownames("GeneID")
Ec_TPM <- readr::read_csv("data/hourglass_data/Ec_abundance.csv") |> 
        # dplyr::mutate(GeneID = stringr::str_remove(gene_id, "\\..$")) |>
        tibble::column_to_rownames("GeneID")
```

```{r}
# check Kryuchkova-Mostacci & Robinson-Rechavi for more functions to add.
specificity <- function(abundance, measure = "tau"){
        abundance_longer <- 
                abundance |>
                duckplyr::as_duckplyr_tibble() |>
                duckplyr::select(1, where(is.numeric)) |>
                tidyr::pivot_longer(!1, names_to = "sample", values_to = "TPM")
        if(measure == "gini"){
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                gini = DescTools::Gini(TPM),
                                .by = GeneID
                        )
        } else if(measure == "tau"){
                n_col <- ncol(abundance) - 1
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                tau = sum(1 - (TPM/max(TPM)))/(n_col-1),
                                .by = GeneID
                        )
        } else if(measure == "H"){
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                H = DescTools::Entropy(TPM/sum(TPM)),
                                .by = GeneID
                        )
        } else if(measure == "rH"){
                # KL divergence from uniform distribution
                n_col <- ncol(abundance) - 1
                x_1 <- 
                breadth_out <- abundance_longer |>
                        duckplyr::mutate(
                                # rH = entropy::KL.empirical(y1 = TPM/sum(TPM), y2 = 1/n_col),
                                y1 = TPM/sum(TPM),
                                y2 = 1/n_col,
                                .by = GeneID
                        ) |>
                        duckplyr::summarise(
                                rH = entropy::KL.empirical(y1, y2),
                                .by = GeneID
                        )
        } else if(measure == "mean"){
                # not really a specificity but useful nonetheless
                breadth_out <- abundance_longer |>
                        duckplyr::summarise(
                                meanTPM = mean(TPM),
                                .by = GeneID
                        )
        } else {
                stop("Specify measure: \"gini\", \"tau\", \"H\", \"rH\", \"mean\"")
        }
        return(breadth_out)
}
```

```{r message=FALSE, warning=FALSE, results='hide'}
Ec_TPM.log2 <- log2(Ec_TPM + 1) |> tibble::rownames_to_column("GeneID")
# Ec_TPM.log2 <- log2(Ec_TPM[,c(7:10)] + 1) |> tibble::rownames_to_column("GeneID")

Ec_tau.log2 <- specificity(Ec_TPM.log2, measure = "tau")
```

Begin with Ec

```{r warning=FALSE, message=FALSE, fig.width=2, fig.height=2.5}
Ec_tau.processed <-
        Ec_tau.log2 |>
        dplyr::mutate(single_copy_og = dplyr::case_when(
                GeneID %in% og_sc.edit.strict$gene_id ~ "yes",
                .default = "no"
        )) 
Ec_tau.processed |>
        ggplot2::ggplot(aes(x = single_copy_og, y = tau, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "1-to-1 orthologue", y = "tau") +
        ggplot2::scale_color_viridis_d()

Ec_tau.processed |>
        ggplot2::ggplot(aes(x = single_copy_og, y = tau, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1),dpi = 300) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "1-to-1 orthologue", y = "tau") +
        ggplot2::scale_color_viridis_d() +
        ggpubr::stat_kruskal_test(label = "p.adj", size = 2.5) +
        EnvStats::stat_n_text(size = 2.5)
# note: not all genes from Jero's dataset overlaps with the ones from the hourglass paper.
```

What about single copy AND conserved signatures?

```{r warning=FALSE, message=FALSE, fig.width=2, fig.height=2.5}
Ec_tau.processed2 <-
        Ec_tau.log2 |>
        dplyr::mutate(single_copy_og = dplyr::case_when(
                GeneID %in% og_sc.edit.strict.conserved$gene_id ~ "yes",
                .default = "no"
        )) 
Ec_tau.processed2 |>
        ggplot2::ggplot(aes(x = single_copy_og, y = tau, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1)) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "conserved 1-to-1 orthologue", y = "tau") +
        ggplot2::scale_color_viridis_d()

Ec_tau.processed2 |>
        ggplot2::ggplot(aes(x = single_copy_og, y = tau, colour = single_copy_og)) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.3, alpha = 0.1),dpi = 300) +
        ggplot2::geom_boxplot(width = 0.5, size = 0.8, outliers = FALSE, colour = "black") +
        ggplot2::theme_classic() +
        ggplot2::theme(legend.position = "none") +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "conserved 1-to-1 orthologue", y = "tau") +
        ggplot2::scale_color_viridis_d() +
        ggpubr::stat_kruskal_test(label = "p.adj", size = 2.5) +
        EnvStats::stat_n_text(size = 2.5)
# note: not all genes from Jero's dataset overlaps with the ones from the hourglass paper.
```

## Combining TAU and Gene age

The question is if there is an interesting subset of specific genes that are expressed in certain contexts.

```{r warning=FALSE, message=FALSE}
Ec_tau.processed.filt <- 
        Ec_tau.processed |>
        dplyr::filter(single_copy_og == "yes" & tau > 0.85)

# rank 7 = Phaeo+Schizo
Ec_ps.processed.filt <- 
        Ec_ps.processed |>
        dplyr::filter(single_copy_og == "yes" & rank >= 7) |>
        dplyr::select(GeneID = gene_id, rank, single_copy_og)
Ec_ps.processed.filt2 <- 
        Ec_ps.processed |>
        dplyr::filter(single_copy_og == "yes") |>
        dplyr::select(GeneID = gene_id, rank, single_copy_og)
```

Let's see if there is any overlap between the two filtered gene sets.

```{r warning=FALSE, message=FALSE}
dplyr::inner_join(Ec_tau.processed.filt, Ec_ps.processed.filt)

Ec_domain <- 
        readr::read_tsv("data/Ectocarpus-sp7.filt.processed.faa.tsv") |>
        dplyr::select(1,12,13) |>
        dplyr::rename(GeneID = 1, domainID = 2, domain = 3) |>
        dplyr::distinct() |>
        dplyr::filter(domainID != "-") |>
        # tidyr::separate_rows(goID, sep = "\\|") |>
        # dplyr::mutate(goID = stringr::str_remove(goID, "(\\(InterPro\\))|(\\(PANTHER\\))")) |>
        dplyr::distinct()

dplyr::inner_join(Ec_tau.processed.filt, Ec_ps.processed.filt) |>
        dplyr::left_join(Ec_domain)
```


```{r warning=FALSE, message=FALSE}
Ec_TPM.log2.hightau <-
        Ec_TPM.log2 |>
        dplyr::filter(GeneID %in% Ec_tau.processed.filt$GeneID) |>
        dplyr::select(GeneID, dplyr::matches("F_|M_"))
```

```{r}
expression_set <- Ec_TPM.log2.hightau |> column_to_rownames(var = "GeneID")
set.seed(123)  # for reproducibility
k <- 2
km_res <- kmeans(expression_set, centers = k)

# Step 5: Add cluster assignment to genes
gene_clusters <- 
        data.frame(
                GeneID = Ec_TPM.log2.hightau$GeneID, 
                kmeans = km_res$cluster)

gene_clusters_set <- 
        gene_clusters |>
        dplyr::mutate(kmeans = factor(kmeans)) |>
        dplyr::left_join(
                Ec_ps.processed.filt2 |>
                        dplyr::mutate(rank = factor(rank))
                ) |> 
        dplyr::select(GeneID, kmeans, phylorank = rank)

sample_set <-
        data.frame(
                sample = colnames(expression_set)
        ) |>
        dplyr::mutate(unicell = dplyr::case_when(
                stringr::str_detect(sample, "spore|gamete") ~ "yes",
                .default = "no"
        )) |>
        dplyr::mutate(sex = dplyr::case_when(
                stringr::str_detect(sample, "M_") ~ "M",
                stringr::str_detect(sample, "F_") ~ "F",
                .default = "none"
        )) |>
        dplyr::mutate(generation = dplyr::case_when(
                stringr::str_detect(sample, "GA_") ~ "GA",
                stringr::str_detect(sample, "PSP_") ~ "PSP",
                .default = "unicell"
        ))
expression_set.filt.pre <- 
        Ec_TPM.log2.hightau |> 
        tidyr::pivot_longer(!GeneID, names_to = "sample") |>
        dplyr::left_join(sample_set) |>
        dplyr::mutate(dev = dplyr::case_when(
                stringr::str_detect(sample, "_early") ~ "early",
                stringr::str_detect(sample, "_imm") ~ "imm",
                stringr::str_detect(sample, "_mat") ~ "mat",
                stringr::str_detect(sample, "_old") ~ "old",
                stringr::str_detect(sample, "_meiospore") ~ "meiospore",
                stringr::str_detect(sample, "_mitospore") ~ "mitospore",
                stringr::str_detect(sample, "_gamete") ~ "gamete",
                .default = NA
        )) |>
        tidyr::unite("stage", c(sex, dev, generation), remove = FALSE) |>
        dplyr::mutate(stage = stringr::str_remove(stage, "_unicell")) |>
        dplyr::summarise(log2TPM = median(value), .by = c(stage, GeneID, unicell, sex, generation,dev))
expression_set.filt <-
        expression_set.filt.pre |>
        dplyr::select(stage, GeneID, log2TPM) |>
        tidyr::pivot_wider(values_from = log2TPM, names_from = stage) |>
        column_to_rownames(var = "GeneID")
```


```{r warning=FALSE, message=FALSE, fig.height=5, fig.width=7}
pheatmap::pheatmap(
        expression_set.filt,
        annotation_row = gene_clusters_set |> column_to_rownames(var = "GeneID"),
        annotation_col = distinct(expression_set.filt.pre, stage, unicell) |> column_to_rownames(var = "stage"),
        cluster_rows = TRUE, 
        cluster_cols = TRUE)#,
        # show_rownames = FALSE,
        # show_colnames = FALSE
```

So the k-means clustering makes sense!

```{r warning=FALSE, message=FALSE, fig.height=4, fig.width=5}
expression_set.filt.pre |>
        dplyr::left_join(gene_clusters_set) |>
        dplyr::mutate(devstage = stringr::str_remove(stage, "M_|F_")) |>
        dplyr::mutate(devstage = stringr::str_remove(devstage, "_")) |>
        dplyr::mutate(devstage = factor(devstage, levels = unique(devstage))) |>
        ggplot2::ggplot(aes(
                x = devstage, y = log2TPM, colour = sex, 
                group = interaction(devstage, sex))) +
        ggplot2::geom_boxplot(colour = "black", outliers = FALSE) +
        ggrastr::rasterise(ggplot2::geom_jitter(width = 0.1, alpha = 0.5), dpi = 300) +
        ggplot2::coord_flip() +
        ggplot2::theme_bw() +
        ggplot2::facet_grid(sex ~ kmeans) +
        ggplot2::labs(x = "developmental stage", y = "median log2(TPM+1)",
                      title = "ortholog set with tau > 0.85") +
        ggplot2::scale_colour_manual(values = c("red", "navy"))
```

```{r warning=FALSE, message=FALSE}
# what is in kmeans 2?
expression_set.filt.pre |>
        dplyr::left_join(gene_clusters_set) |>
        dplyr::distinct(GeneID, kmeans) |>
        dplyr::filter(kmeans == 2) |>
        dplyr::left_join(Ec_domain)
expression_set.filt.pre |>
        dplyr::left_join(gene_clusters_set) |>
        dplyr::distinct(GeneID, kmeans) |>
        dplyr::filter(kmeans == 2) |>
        dplyr::left_join(Ec_domain) |>
        dplyr::filter(stringr::str_detect(domain, "AAA|ATPase")) |>
        dplyr::distinct(GeneID)
```

There are two genes with ATPase-related domains, `Ec-02_005230.1` and `Ec-11_003070.1`. I am unsure what to make of this.

## Signature enrichment

```{r warning=FALSE, message=FALSE}
Ec_odds.in <- 
        Ec_ps.processed |>
        dplyr::count(Signature.Female, single_copy_og) |>
        tidyr::pivot_wider(
                names_from = single_copy_og,
                values_from = n,
                values_fill = 0
                )
```


```{r warning=FALSE, message=FALSE}
# Total counts
total_yes <- sum(Ec_odds.in$yes)
total_no  <- sum(Ec_odds.in$no)

Ec_odds.out <- 
        Ec_odds.in |>
        dplyr::mutate(
                A = yes,
                B = no,
                C = total_yes - yes,
                D = total_no - no,
                odds_ratio = (A * D) / (B * C),
                # Optional: Fisher's exact test for significance
                p_value = fisher.test(matrix(c(A, B, C, D), nrow = 2))$p.value,
                .by = Signature.Female
                ) |>
        dplyr::mutate(p_adj = p.adjust(p_value, method = "BH"))

Ec_odds.out |>
        dplyr::mutate(p_adj.view = dplyr::case_when(
                p_adj > 0.05 ~ "n.s.",
                p_adj < 0.001 ~ "***",
                p_adj < 0.01 ~ "**",
                p_adj < 0.05 ~ "*",
                .default = NA
        )) |>
        dplyr::select(1:3, odds_ratio, p_adj, p_adj.view)
```

```{r warning=FALSE, message=FALSE, fig.height=3, fig.width=2.5}
color_table <-
        tibble::tibble(
                Signature = paste0('S',1:16),
                Colour = c("#a31d21", "#e43a30", "#f36f57", "#f8a68d", "#fddfd1", "#1b733b",
            "#2daf66", "#6ec283", "#aed9b1", "#9392c8", "#bfbfdf", "#e9eef8",
            "#2a74bb", "#61abd7", "#afd1e5", "#999999")
        )
Ec_odds.out |>
        dplyr::mutate(Signature.Female = factor(Signature.Female, levels = paste0('S',1:16))) |>
        ggplot2::ggplot(aes(x = Signature.Female, 
                            y = log2(odds_ratio), 
                            fill = Signature.Female, 
                            group = 1)) +
        ggplot2::geom_hline(yintercept = 0, colour = "grey50", linetype = 2) +
        ggplot2::geom_line(colour = "black", linetype = 2) +
        ggplot2::geom_point(size = 2, shape = 21, colour = "black") +
        ggplot2::theme_bw() +
        ggplot2::scale_fill_manual(values = rev(color_table$Colour)) +
        ggplot2::theme(legend.position = "none") +
        ggplot2::coord_flip() +
        # ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "Signature", y = "log2(odds ratio)", 
                      title = "Ectocarpus",
                      subtitle = "Enrichment in 1to1 orthologs")
```

```{r warning=FALSE, message=FALSE, fig.height=3, fig.width=4}
Ec_odds.out |>
        dplyr::mutate(Signature.Female = factor(Signature.Female, levels = paste0('S',1:16))) |>
        ggplot2::ggplot(aes(x = Signature.Female, 
                            y = log2(odds_ratio), 
                            fill = Signature.Female, 
                            group = 1)) +
        ggplot2::geom_hline(yintercept = 0, colour = "grey50", linetype = 2) +
        ggplot2::geom_line(colour = "black", linetype = 2) +
        ggplot2::geom_point(shape = 21, colour = "black", aes(size = -log10(p_adj))) +
        ggplot2::theme_classic() +
        ggplot2::scale_fill_manual(values = rev(color_table$Colour)) +
        ggplot2::guides(
                size = guide_legend(position = "right"),
                fill = "none"
                ) +
        ggplot2::coord_flip() +
        # ggplot2::scale_x_discrete(limits=rev) +
        ggplot2::scale_y_continuous(breaks = scales::breaks_pretty()) +
        ggplot2::labs(x = "Signature", y = "log2(odds ratio)", 
                      title = "Ectocarpus",
                      subtitle = "Enrichment in 1to1 orthologs")
```
